---
title: "FIN 452 — Quantitative Trading Strategy"
subtitle: "Clustering + Kalman Filter | S&P 500 Select Sector ETFs"
author: "Ken Sucento"
date: "April 7, 2026"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    theme: flatly
    number-sections: true
---

```{r}
#| echo: false
#| warning: false
#| message: false
library(tidyverse)
library(tidyquant)
library(tidymodels)
library(rlang)
library(purrr)
library(lubridate)
library(plotly)
library(patchwork)
library(gt)
library(cluster)
library(factoextra)
library(tseries)
library(KFAS)
library(broom)
library(slider)
library(TTR)
library(PerformanceAnalytics)
library(RTL)
```

## Mental Model - How the Strategy Came to Life

The strategy answers one question: *"Can two sector ETFs that historically move together be traded profitably when their spread temporarily diverges?"* The answer requires four sequential filters each one narrows the universe from many candidates to one actionable pair and a concrete trading rule. The diagram below shows the full lifecycle from hypothesis to evaluation.

```{=html}
<iframe src="qtmm.pdf" width="100%" height="750px" style="border:1px solid #ddd;border-radius:6px;"></iframe>
```

## Data

### ETF Universe

The strategy universe consists of eleven S&P 500 Select Sector SPDR ETFs plus the benchmark (SPY). Two ETFs have limited history and are **excluded from the clustering model**:

-   **XLC** (Communication Services) — available from June 2018 only; only \~6.5 years of training data
-   **XLRE** (Real Estate) — available from October 2015 only; only \~9 years of training data

Both are retained in the full data pull for possible use in later pipeline steps.

```{r}
#| echo: false
#| warning: false
#| message: false
#| cache: true

tickers <- c("SPY", "XLB", "XLC", "XLE", "XLF", "XLI",
             "XLK", "XLP", "XLRE", "XLV", "XLU", "XLY")

df <- tidyquant::tq_get(
  tickers,
  get  = "stock.prices",
  from = "2000-06-01",
  to   = "2026-03-31"
)
```

## Background Setup

```{r}
#| include: false

df.train <- df %>% dplyr::filter(date <= as.Date("2024-12-31"))
df.test  <- df %>% dplyr::filter(date >= as.Date("2025-01-01"))

etf.ret.train <- df.train %>%
  dplyr::group_by(symbol) %>%
  dplyr::mutate(ret = log(adjusted / dplyr::lag(adjusted))) %>%
  tidyr::drop_na(ret) %>%
  dplyr::ungroup()

etf.ret.test <- df.test %>%
  dplyr::group_by(symbol) %>%
  dplyr::mutate(ret = log(adjusted / dplyr::lag(adjusted))) %>%
  tidyr::drop_na(ret) %>%
  dplyr::ungroup()
```

## Step 1 - Clustering

Before we can test any pair for mean-reversion, we need to narrow 10 sector ETFs down to pairs that have a structural economic reason to move together.

Clustering on daily log returns does that. We use log returns rather than prices because prices drift upward over 24 years, which makes the raw distance between two price series meaningless as a similarity measure. Log returns strip away that drift and leave only co-movement, which is the actual signal we care about. **We also cluster on returns alone rather than mixing in macro variables like interest rates, oil prices, or GDP growth. Those factors are already embedded in how ETFs move every day.**

If XLE is sensitive to oil, that sensitivity shows up in its realized returns relative to other sectors. Adding macro variables explicitly would double-count information that is already in the data, while introducing noise from slow-moving, non-stationary series that distort the distance calculation. XLC and XLRE are excluded because both have under 10 years of training history, which is not enough to estimate stable long-run relationships.

```{r}
#| echo: false
#| warning: false
#| message: false

# --- Data prep ---
tickers.cluster <- c("SPY", "XLB", "XLE", "XLF", "XLI",
                     "XLK", "XLP", "XLV", "XLU", "XLY")

etf.wide.train <- etf.ret.train %>%
  dplyr::filter(symbol %in% tickers.cluster) %>%
  tidyr::pivot_wider(
    id_cols     = date,
    names_from  = symbol,
    values_from = ret
  ) %>%
  tidyr::drop_na()

recipe.cluster <- recipes::recipe(~ ., data = etf.wide.train) %>%
  recipes::step_rm(date) %>%
  recipes::step_normalize(recipes::all_numeric()) %>%
  recipes::prep(training = etf.wide.train)

etf.baked  <- recipes::bake(recipe.cluster, new_data = etf.wide.train)
etf.scaled <- base::as.matrix(etf.baked)
etf.t      <- base::t(etf.scaled)
dist.mat   <- stats::dist(etf.t)

# --- WSS for elbow ---
wss <- purrr::map_dbl(2:8, function(k) {
  km <- stats::kmeans(etf.t, centers = k, nstart = 25)
  km$tot.withinss
})
df.elbow <- dplyr::tibble(k = 2:8, wss = wss)

# --- K-Means (k = 3) ---
set.seed(452)
km.result <- stats::kmeans(etf.t, centers = 3, nstart = 25)

def.cluster <- km.result$cluster[["XLU"]]
cyc.cluster <- km.result$cluster[["XLE"]]
grw.cluster <- km.result$cluster[["XLK"]]

df.clusters <- dplyr::tibble(
  symbol  = base::names(km.result$cluster),
  cluster = km.result$cluster
) %>%
  dplyr::mutate(
    cluster.label = dplyr::case_when(
      cluster == def.cluster ~ "Defensive",
      cluster == cyc.cluster ~ "Value / Cyclical",
      cluster == grw.cluster ~ "Growth",
      TRUE ~ paste0("Cluster ", cluster)
    ),
    sector = dplyr::case_when(
      symbol == "SPY"  ~ "S&P 500 Benchmark",
      symbol == "XLB"  ~ "Materials",
      symbol == "XLE"  ~ "Energy",
      symbol == "XLF"  ~ "Financials",
      symbol == "XLI"  ~ "Industrials",
      symbol == "XLK"  ~ "Technology",
      symbol == "XLP"  ~ "Consumer Staples",
      symbol == "XLV"  ~ "Health Care",
      symbol == "XLU"  ~ "Utilities",
      symbol == "XLY"  ~ "Consumer Discretionary",
      TRUE ~ symbol
    )
  ) %>%
  dplyr::arrange(cluster.label, symbol)

# --- Hierarchical (Ward's) ---
hc.result <- stats::hclust(dist.mat, method = "ward.D2")

# --- Silhouette ---
sil     <- cluster::silhouette(km.result$cluster, dist.mat)
sil.avg <- base::mean(sil[, 3])

# --- Stability: pre vs post 2020 ---
etf.wide.pre2020 <- etf.ret.train %>%
  dplyr::filter(symbol %in% tickers.cluster, date < as.Date("2020-01-01")) %>%
  tidyr::pivot_wider(id_cols = date, names_from = symbol, values_from = ret) %>%
  tidyr::drop_na()

recipe.pre2020 <- recipes::recipe(~ ., data = etf.wide.pre2020) %>%
  recipes::step_rm(date) %>%
  recipes::step_normalize(recipes::all_numeric()) %>%
  recipes::prep(training = etf.wide.pre2020)

etf.pre2020 <- base::t(base::as.matrix(
  recipes::bake(recipe.pre2020, new_data = etf.wide.pre2020)
))

set.seed(452)
km.pre2020 <- stats::kmeans(etf.pre2020, centers = 3, nstart = 25)

etf.wide.post2020 <- etf.ret.train %>%
  dplyr::filter(symbol %in% tickers.cluster, date >= as.Date("2020-01-01")) %>%
  tidyr::pivot_wider(id_cols = date, names_from = symbol, values_from = ret) %>%
  tidyr::drop_na()

recipe.post2020 <- recipes::recipe(~ ., data = etf.wide.post2020) %>%
  recipes::step_rm(date) %>%
  recipes::step_normalize(recipes::all_numeric()) %>%
  recipes::prep(training = etf.wide.post2020)

etf.post2020 <- base::t(base::as.matrix(
  recipes::bake(recipe.post2020, new_data = etf.wide.post2020)
))

set.seed(452)
km.post2020 <- stats::kmeans(etf.post2020, centers = 3, nstart = 25)

def.pre  <- km.pre2020$cluster[["XLU"]]
cyc.pre  <- km.pre2020$cluster[["XLE"]]
grw.pre  <- km.pre2020$cluster[["XLK"]]
def.post <- km.post2020$cluster[["XLU"]]
cyc.post <- km.post2020$cluster[["XLE"]]
grw.post <- km.post2020$cluster[["XLK"]]

df.stability <- dplyr::tibble(
  symbol    = base::names(km.pre2020$cluster),
  pre.2020  = dplyr::case_when(
    km.pre2020$cluster == def.pre  ~ "Defensive",
    km.pre2020$cluster == cyc.pre  ~ "Value / Cyclical",
    km.pre2020$cluster == grw.pre  ~ "Growth",
    TRUE ~ "Other"
  ),
  post.2020 = dplyr::case_when(
    km.post2020$cluster == def.post  ~ "Defensive",
    km.post2020$cluster == cyc.post  ~ "Value / Cyclical",
    km.post2020$cluster == grw.post  ~ "Growth",
    TRUE ~ "Other"
  )
) %>%
  dplyr::mutate(stable = dplyr::if_else(pre.2020 == post.2020, "Yes", "No"))
```

### Cluster Assignments

Three groups emerge that map onto how institutional investors actually think about sector rotation.

-   Defensive sectors (XLP, XLU) hold their value in downturns and carry low market beta.

-   Value/Cyclical sectors (XLE, XLF) are driven by commodity prices and interest rate regimes, not the broad market.

-   The Growth cluster captures the remaining sectors that are technology-heavy and tightly tied to SPY.

**Both the elbow plot and the dendrogram support k = 3.** The elbow has no sharp kink, but that is expected: every sector ETF carries heavy SPY exposure, which compresses the distance between clusters.

```{r}
#| echo: false
#| warning: false
#| message: false

p.elbow <- df.elbow %>%
  ggplot2::ggplot(ggplot2::aes(x = k, y = wss)) +
  ggplot2::geom_line(col = "blue") +
  ggplot2::geom_point(col = "blue", size = 3) +
  ggplot2::geom_vline(xintercept = 3, col = "red", linetype = "dashed") +
  ggplot2::scale_x_continuous(breaks = 2:8) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title    = "Elbow Method — K-Means WSS by Number of Clusters",
    subtitle = "Red dashed line at k = 3 | Training data only: Jun 2000 – Dec 2024",
    x        = "Number of Clusters (k)",
    y        = "Total Within-Cluster SS"
  )

p.dend <- factoextra::fviz_dend(
  hc.result,
  k         = 3,
  rect      = TRUE,
  rect_fill = TRUE,
  ggtheme   = ggplot2::theme_minimal(),
  main      = "Dendrogram — Ward's Method (k = 3)",
  sub       = "Training period: Jun 2000 – Dec 2024",
  xlab      = "",
  ylab      = "Height"
)

p.elbow
p.dend

df.clusters %>%
  dplyr::select(symbol, sector, cluster.label) %>%
  dplyr::rename(
    Ticker  = symbol,
    Sector  = sector,
    Cluster = cluster.label
  ) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "K-Means Cluster Assignments (k = 3)",
    subtitle = "Fit on training data Jun 2000 – Dec 2024 | Seed = 452"
  ) %>%
  gt::tab_row_group(label = "Growth",          rows = Cluster == "Growth") %>%
  gt::tab_row_group(label = "Value / Cyclical", rows = Cluster == "Value / Cyclical") %>%
  gt::tab_row_group(label = "Defensive",        rows = Cluster == "Defensive") %>%
  gt::tab_footnote(
    footnote  = "XLC (Communication Services) and XLRE (Real Estate) excluded due to insufficient training history.",
    locations = gt::cells_title(groups = "subtitle")
  )
```

### Silhouette Validation

The average silhouette score is approximately 0.16, which is low but not a red flag. It reflects the structure of the data rather than a problem with the clustering. **Because every sector ETF carries beta close to 1.0 against SPY, the algorithm is looking for sub-structure inside a universe where everything is already highly correlated. Tight, well-separated clusters are not achievable here.**

The purpose of clustering is to nominate economically plausible pairs, not to find perfect separation. The statistical validation comes in Step 2 through cointegration testing.

```{r}
#| echo: false
#| warning: false
#| message: false

p.sil <- factoextra::fviz_silhouette(
  sil,
  ggtheme       = ggplot2::theme_minimal(),
  print.summary = FALSE
) +
  ggplot2::labs(
    title    = paste0(
      "Silhouette Plot — K-Means (k = 3) | Avg Score = ",
      base::round(sil.avg, 3)
    ),
    subtitle = "Low score expected: sector ETFs share systematic market exposure by construction"
  )

p.sil

dplyr::tibble(
  Metric = c("Average Silhouette Score", "Cluster Quality"),
  Value  = c(
    base::as.character(base::round(sil.avg, 3)),
    dplyr::case_when(
      sil.avg >= 0.50 ~ "Strong (>= 0.5) — proceed",
      sil.avg >= 0.30 ~ "Moderate (0.3 – 0.5) — acceptable",
      TRUE            ~ "Weak (< 0.3) — revisit k or features"
    )
  )
) %>%
  gt::gt() %>%
  gt::tab_header(title = "K-Means Cluster Validation — Silhouette") %>%
  gt::tab_style(
    style     = gt::cell_text(color = "red", weight = "bold"),
    locations = gt::cells_body(
      columns = Value,
      rows    = Metric == "Cluster Quality" & grepl("Weak", Value)
    )
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(color = "blue", weight = "bold"),
    locations = gt::cells_body(
      columns = Value,
      rows    = Metric == "Cluster Quality" & grepl("Strong|Moderate", Value)
    )
  )
```

------------------------------------------------------------------------

## Step 2 - Cointegration Testing

Clustering tells us which sectors have historically moved together, but economic similarity is not the same as a tradeable relationship. Two ETFs can be highly correlated over 24 years and still have a spread that drifts permanently in one direction. A persistent drift means every trade betting on reversion compounds losses over time. **We need statistical proof that the spread is bounded and mean-reverting before committing capital to any pair.**

That proof comes from cointegration testing. We follow the Engle-Granger two-step procedure as implemented by Yang, Huang and Chen (2023).

-   The first step runs an OLS regression of log(P_A) on log(P_B) to estimate the hedge ratio beta. This beta determines how many units of ETF B to hold short against one long unit of ETF A so the resulting spread has the smallest possible long-run drift.

-   The second step runs an Augmented Dickey-Fuller test on those OLS residuals, which are the spread itself. The ADF null hypothesis is that the spread has a unit root, meaning it can wander arbitrarily far from its mean and never return. Rejecting that null at p \< 0.05 confirms the spread is stationary and mean-reverting. The pair is cointegrated and proceeds to Step 3.

Any pair that fails is dropped entirely, because trading a non-stationary spread has no theoretical bound on the loss. Every within-cluster pair from Step 1 is tested.

```{r}
#| echo: false
#| warning: false
#| message: false

# Log-price wide matrices — CLOSE (used for cointegration, OU, Kalman, and CC return)
log.prices.wide.train <- df.train %>%
  dplyr::filter(symbol %in% tickers.cluster) %>%
  dplyr::select(date, symbol, adjusted) %>%
  tidyr::pivot_wider(names_from = symbol, values_from = adjusted) %>%
  dplyr::mutate(dplyr::across(where(is.numeric), log)) %>%
  tidyr::drop_na()

log.prices.wide.test <- df.test %>%
  dplyr::filter(symbol %in% tickers.cluster) %>%
  dplyr::select(date, symbol, adjusted) %>%
  tidyr::pivot_wider(names_from = symbol, values_from = adjusted) %>%
  dplyr::mutate(dplyr::across(where(is.numeric), log)) %>%
  tidyr::drop_na()

# Log-price wide matrices — OPEN (adjusted for splits/dividends via adj ratio)
# Adj ratio = adjusted / close; apply to open: open.adj = open * (adjusted / close)
log.open.wide.train <- df.train %>%
  dplyr::filter(symbol %in% tickers.cluster) %>%
  dplyr::mutate(open.adj = open * (adjusted / close)) %>%
  dplyr::select(date, symbol, open.adj) %>%
  tidyr::pivot_wider(names_from = symbol, values_from = open.adj) %>%
  dplyr::mutate(dplyr::across(where(is.numeric), log)) %>%
  tidyr::drop_na()

log.open.wide.test <- df.test %>%
  dplyr::filter(symbol %in% tickers.cluster) %>%
  dplyr::mutate(open.adj = open * (adjusted / close)) %>%
  dplyr::select(date, symbol, open.adj) %>%
  tidyr::pivot_wider(names_from = symbol, values_from = open.adj) %>%
  dplyr::mutate(dplyr::across(where(is.numeric), log)) %>%
  tidyr::drop_na()
```

### Candidate Pair Generation

All within-cluster pairs are generated from the K-Means assignments. Pairs from different clusters are excluded as they have no structural economic relationship.

```{r}
#| echo: false
#| warning: false
#| message: false

# Generate all within-cluster pairs (no duplicates, no self-pairs)
df.pairs <- df.clusters %>%
  dplyr::rename(symbolA = symbol, clusterA = cluster.label) %>%
  dplyr::inner_join(
    df.clusters %>% dplyr::rename(symbolB = symbol, clusterB = cluster.label),
    by           = c("clusterA" = "clusterB"),
    relationship = "many-to-many"
  ) %>%
  dplyr::filter(symbolA < symbolB) %>%
  dplyr::select(symbolA, symbolB, cluster = clusterA)
```

### ADF Test - Engle-Granger Cointegration

For each candidate pair (A, B), OLS is run on log prices: `log(P_A) ~ log(P_B)`. The ADF test on the residuals determines whether the spread is stationary. Pairs with p-value \< 0.05 are cointegrated and proceed.

```{r}
#| echo: false
#| warning: false
#| message: false

df.coint <- df.pairs %>%
  dplyr::mutate(
    beta.ols = purrr::map2_dbl(symbolA, symbolB, function(a, b) {
      fit <- stats::lm(
        log.prices.wide.train[[a]] ~ log.prices.wide.train[[b]]
      )
      broom::tidy(fit)$estimate[2]
    }),
    adf.stat = purrr::map2_dbl(symbolA, symbolB, function(a, b) {
      fit   <- stats::lm(
        log.prices.wide.train[[a]] ~ log.prices.wide.train[[b]]
      )
      resid <- stats::resid(fit)
      base::suppressWarnings(tseries::adf.test(resid))$statistic
    }),
    adf.pval = purrr::map2_dbl(symbolA, symbolB, function(a, b) {
      fit   <- stats::lm(
        log.prices.wide.train[[a]] ~ log.prices.wide.train[[b]]
      )
      resid <- stats::resid(fit)
      base::suppressWarnings(tseries::adf.test(resid))$p.value
    }),
    cointegrated = dplyr::if_else(adf.pval < 0.05, "Yes", "No")
  )
```

```{r}
#| echo: false
#| warning: false
#| message: false

df.coint %>%
  dplyr::arrange(adf.pval) %>%
  dplyr::select(symbolA, symbolB, cluster, beta.ols, adf.stat, adf.pval, cointegrated) %>%
  dplyr::rename(
    `ETF A`        = symbolA,
    `ETF B`        = symbolB,
    Cluster        = cluster,
    `OLS Beta`     = beta.ols,
    `ADF Stat`     = adf.stat,
    `ADF p-value`  = adf.pval,
    Cointegrated   = cointegrated
  ) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Cointegration Test Results — All Candidate Pairs",
    subtitle = "Engle-Granger ADF test on OLS residuals | Training period: Jun 2000 – Dec 2024"
  ) %>%
  gt::fmt_number(columns = c(`OLS Beta`, `ADF Stat`, `ADF p-value`), decimals = 4) %>%
  gt::tab_style(
    style     = gt::cell_text(color = "blue", weight = "bold"),
    locations = gt::cells_body(
      columns = Cointegrated,
      rows    = Cointegrated == "Yes"
    )
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(color = "red"),
    locations = gt::cells_body(
      columns = Cointegrated,
      rows    = Cointegrated == "No"
    )
  ) %>%
  gt::tab_footnote(
    footnote  = "Null hypothesis: series has unit root (non-stationary). Reject at p < 0.05 → spread is stationary → pair is cointegrated.",
    locations = gt::cells_column_labels(columns = `ADF p-value`)
  )
```

### Cointegration Results

The test narrows the 22 within-cluster candidate pairs down to a smaller set of cointegrated survivors.

Most of the failures come from the Defensive and Value/Cyclical clusters, where idiosyncratic drivers such as recession demand and commodity price cycles make it difficult for spreads to stay bounded over 24 years. **The pairs that survive come from the Growth cluster, where shared technology exposure and market-cap concentration create tighter structural linkages that persist across full business cycles.**

------------------------------------------------------------------------

## Step 3 - OU Half-Life

Cointegration confirms that a mean-reverting spread exists, but it says nothing about how quickly it reverts.

A pair with a half-life of 240 days is statistically cointegrated and practically useless in the same breath. Holding a position for nearly a full year ties up capital, accumulates financing costs, and gives the underlying relationship more time to structurally break down mid-trade. We need to rank the surviving pairs by reversion speed and select the one that converts a statistical finding into an actionable holding period.

We used the Ornstein-Uhlenbeck process to models the spread as a continuous mean-reverting system. The reversion speed is estimated by regressing the daily change in the spread on the prior day's spread level. The coefficient on the lagged spread, called theta, captures the pull-back force toward the mean. A more negative theta means stronger and faster reversion. The half-life is then -ln(2) / theta, which converts that coefficient into a number of days.

Any pair where theta is non-negative has no mean-reverting pull and is discarded. From the remaining **pairs, a ceiling of 252 trading days is applied.** This is more generous than the 60-day cutoff commonly used in individual equity pairs trading, because sector ETFs are diversified portfolios with idiosyncratic risk already averaged out. Their residual mean-reversion operates on longer timescales by construction.

```{r}
#| echo: false
#| warning: false
#| message: false

df.valid <- df.coint %>%
  dplyr::filter(cointegrated == "Yes") %>%
  dplyr::mutate(
    half.life = purrr::pmap_dbl(
      list(symbolA, symbolB, beta.ols),
      function(a, b, hr) {
        spread <- log.prices.wide.train[[a]] - hr * log.prices.wide.train[[b]]
        df.ou  <- dplyr::tibble(
          spread     = spread,
          spread.lag = dplyr::lag(spread),
          delta      = spread - spread.lag
        ) %>% tidyr::drop_na()
        fit.ou <- stats::lm(delta ~ spread.lag, data = df.ou)
        theta  <- broom::tidy(fit.ou)$estimate[2]
        -base::log(2) / theta
      }
    ),
    tradeable = dplyr::if_else(
      half.life > 0 & half.life <= 252, "Yes", "No"
    )
  ) %>%
  dplyr::arrange(half.life)
```

```{r}
#| echo: false
#| warning: false
#| message: false

df.valid %>%
  dplyr::select(symbolA, symbolB, cluster, beta.ols, adf.pval, half.life, tradeable) %>%
  dplyr::rename(
    `ETF A`       = symbolA,
    `ETF B`       = symbolB,
    Cluster       = cluster,
    `OLS Beta`    = beta.ols,
    `ADF p-value` = adf.pval,
    `Half-Life (days)` = half.life,
    Tradeable     = tradeable
  ) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "OU Half-Life — Cointegrated Pairs",
    subtitle = "Short half-life = faster mean reversion = more tradeable | Discard HL > 252 days or θ ≥ 0"
  ) %>%
  gt::fmt_number(
    columns  = c(`OLS Beta`, `ADF p-value`, `Half-Life (days)`),
    decimals = 3
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(color = "blue", weight = "bold"),
    locations = gt::cells_body(
      columns = Tradeable,
      rows    = Tradeable == "Yes"
    )
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(color = "red"),
    locations = gt::cells_body(
      columns = Tradeable,
      rows    = Tradeable == "No"
    )
  )
```

```{r}
#| include: false

# All tradeable pairs — traded simultaneously with equal weight
pairs.tradeable <- df.valid %>%
  dplyr::filter(tradeable == "Yes") %>%
  dplyr::arrange(half.life)

# Primary pair (shortest half-life) — retained as reference label for CC/CO/OC diagnostics
tickerA <- pairs.tradeable$symbolA[1]
tickerB <- pairs.tradeable$symbolB[1]
```

All `r nrow(pairs.tradeable)` tradeable pairs proceed to the strategy layer and are traded simultaneously with equal weight allocation, rebalanced to equal weight at the end of each calendar month.

### Half-Life Results

All tradeable pairs come from the Growth cluster. The Defensive and Value/Cyclical pairs were eliminated in Step 2, which confirms that structural economic linkage does not guarantee a tradeable spread. Growth ETFs dominate because shared technology exposure and large-cap concentration keep their spreads from drifting permanently across business cycles.

The fastest-reverting pair is XLV/XLY (Health Care vs Consumer Discretionary). Both are large-cap domestic sectors with similar overall market sensitivity, but they respond to opposite ends of the consumer cycle. Health Care spending is non-discretionary and holds up in downturns. Consumer Discretionary contracts when confidence falls. Because neither sector permanently outpaces the other over a full business cycle, the spread oscillates around a stable mean. Rather than concentrating the entire strategy in one pair, all tradeable pairs are traded simultaneously. Each pair contributes an independent source of mean-reversion alpha, and monthly rebalancing to equal weight systematically sells pairs that have outperformed and buys those that have underperformed, compounding the mean-reversion effect at the portfolio level.

------------------------------------------------------------------------

## Step 4 - Kalman Filter and Strategy

### Strategy Function

Cointegration confirmed that the spread between two ETFs is stationary, it cannot drift permanently. But to actually trade it, we need to know how many units of ETF B to short against one long unit of ETF A at any given time. That ratio is the hedge ratio β.

The naive approach is to reuse the OLS beta from Step 2. The problem: that beta is a single point estimate fit on 24 years of price history. Sector relationships shift across business cycles. The same two ETFs that moved dollar-for-dollar in 2005 may have drifted to a 1.2:1 relationship by 2015 as their compositions and market roles evolved. A static hedge ratio will progressively mismatch the spread, generating signals against a spread that no longer means what it used to.

**The Kalman filter solves this.** Introduced by Kalman (1960) and applied to pairs trading by Avellaneda & Lee (2008) and Yang, Huang & Chen (2023), it is a recursive Bayesian estimator that updates β_t every single day.

Mechanically :

-   It treats the true hedge ratio as a hidden state that evolves over time. Each new price observation updates the estimate by blending the prior day's value with a correction term scaled by the Kalman gain K_t. The gain is governed by the ratio Q/H, process noise to observation noise. High Q/H means the hedge ratio is volatile, so trust the new data more. Low Q/H means the hedge ratio is stable, so stay close to the prior.

-   The calibration follows Yang, Huang & Chen (2023): Q = α² / (1 − α) × H, where α = 2 / (n.ma + 1). This ties the filter's adaptation speed directly to a target moving-average window n.ma. A larger n.ma produces a smoother, slower-adapting β_t — appropriate for sector pairs that mean-revert over months. The optimal n.ma is determined by the grid search in Section 8.7.

-   Once β_t is known, the spread is: spread_t = log(P_A,t) − β_t × log(P_B,t). The z-score of this spread against its 20-day rolling mean and standard deviation is the trade signal. Z \> +z.entry means the spread is stretched, short it. Z \< −z.entry means the spread is compressed, long it. The position closes when z returns to zero, or is stopped at ±z.stop.

The full strategy logic is encapsulated in two functions. `strategy_pair()` runs the Kalman filter and signal generation for a single pair. `portfolio_pairs()` wraps it across all tradeable pairs, aggregates their daily returns with equal weight allocation, and rebalances to equal weight at the end of each calendar month.

```{r}
#| echo: false
#| warning: false
#| message: false

strategy_pair <- function(prices.wide,
                          tickerA,
                          tickerB,
                          n.ma       = 60,
                          z.entry    = 2,
                          z.stop     = 3,
                          open.wide  = NULL) {

  alpha <- 2 / (n.ma + 1)
  q     <- alpha^2 / (1 - alpha)

  vec.A  <- prices.wide[[tickerA]]
  vec.B  <- prices.wide[[tickerB]]
  n.obs  <- base::length(vec.A)

  H.est  <- stats::var(vec.A - vec.B, na.rm = TRUE)
  Q.est  <- q * H.est

  # Kalman filter: state = hedge ratio beta_t
  beta_t <- base::numeric(n.obs)
  P.filt <- H.est

  for (i in base::seq_len(n.obs)) {
    P.pred    <- P.filt + Q.est
    Bi        <- vec.B[i]
    K         <- P.pred * Bi / (Bi^2 * P.pred + H.est)
    beta.pred <- if (i == 1L) 1 else beta_t[i - 1L]
    beta_t[i] <- beta.pred + K * (vec.A[i] - Bi * beta.pred)
    P.filt    <- (1 - K * Bi) * P.pred
  }

  result <- prices.wide %>%
    dplyr::mutate(
      beta_t  = beta_t,
      spread  = .data[[tickerA]] - beta_t * .data[[tickerB]],
      z.mean  = slider::slide_dbl(spread, mean,       .before = 19, .after = 0, .complete = FALSE),
      z.sd    = slider::slide_dbl(spread, stats::sd,  .before = 19, .after = 0, .complete = FALSE),
      z.score = (spread - z.mean) / z.sd,
      # Close-to-close log returns per leg
      ret.A   = .data[[tickerA]] - dplyr::lag(.data[[tickerA]]),
      ret.B   = .data[[tickerB]] - dplyr::lag(.data[[tickerB]]),
      signal  = dplyr::case_when(
        base::abs(z.score) > z.stop ~ 0L,
        z.score >  z.entry           ~ -1L,
        z.score < -z.entry           ~ 1L,
        TRUE                         ~ 0L
      ),
      trade   = tidyr::replace_na(
        dplyr::lag(signal) - dplyr::lag(signal, n = 2L), 0L
      ),
      pos          = base::cumsum(trade),
      prev.pos     = dplyr::lag(pos, default = 0L),
      # Close → Close spread return
      spread.ret.cc = ret.A - beta_t * ret.B
    )

  if (!base::is.null(open.wide)) {
    # Compute intraday (OC) and overnight (CO) spread returns
    result <- result %>%
      dplyr::mutate(
        open.A        = open.wide[[tickerA]],
        open.B        = open.wide[[tickerB]],
        ret.co.A      = open.A - dplyr::lag(.data[[tickerA]]),
        ret.co.B      = open.B - dplyr::lag(.data[[tickerB]]),
        spread.ret.co = ret.co.A - beta_t * ret.co.B,
        ret.oc.A      = .data[[tickerA]] - open.A,
        ret.oc.B      = .data[[tickerB]] - open.B,
        spread.ret.oc = ret.oc.A - beta_t * ret.oc.B
      ) %>%
      dplyr::mutate(
        # Entry day (prev flat → in trade): fill at open → earn OC
        # Exit day  (in trade → flat):      exit at open → earn CO on prior leg
        # Hold / flip:                       earn full CC
        ret = dplyr::case_when(
          prev.pos == 0L & pos != 0L ~ pos      * tidyr::replace_na(spread.ret.oc, 0),
          pos == 0L & prev.pos != 0L ~ prev.pos * tidyr::replace_na(spread.ret.co, 0),
          TRUE                       ~ pos      * tidyr::replace_na(spread.ret.cc, 0)
        ),
        cumeq    = base::cumprod(1 + ret),
        ret.co   = pos * tidyr::replace_na(spread.ret.co, 0),
        cumeq.co = base::cumprod(1 + ret.co),
        ret.oc   = pos * tidyr::replace_na(spread.ret.oc, 0),
        cumeq.oc = base::cumprod(1 + ret.oc)
      )
  } else {
    result <- result %>%
      dplyr::mutate(
        ret   = pos * tidyr::replace_na(spread.ret.cc, 0),
        cumeq = base::cumprod(1 + ret)
      )
  }

  result
}
```

```{r}
#| include: false

# portfolio_pairs(): runs strategy_pair() for every row of pairs.df,
# aggregates to an equal-weight portfolio with monthly rebalancing
portfolio_pairs <- function(pairs.df,
                            prices.wide,
                            open.wide  = NULL,
                            n.ma       = 60,
                            z.entry    = 2,
                            z.stop     = 3) {

  # Run strategy_pair for each pair; keep only date + daily CC return
  pair.results <- purrr::pmap(
    list(pairs.df$symbolA, pairs.df$symbolB),
    function(a, b) {
      strategy_pair(
        prices.wide = prices.wide,
        tickerA     = a,
        tickerB     = b,
        n.ma        = n.ma,
        z.entry     = z.entry,
        z.stop      = z.stop,
        open.wide   = open.wide   # pass through for realistic execution timing
      ) %>%
        dplyr::select(date, ret) %>%
        dplyr::rename(!!paste0(a, "_", b) := ret)
    }
  )

  # Align all pair return streams on the same dates
  df.wide   <- purrr::reduce(pair.results, dplyr::inner_join, by = "date") %>%
    tidyr::drop_na()
  pair.cols <- base::setdiff(base::names(df.wide), "date")
  n.p       <- base::length(pair.cols)

  # Equal-weight allocation tracker (normalised to start at 1)
  allocs             <- base::rep(1 / n.p, n.p)
  base::names(allocs) <- pair.cols

  n.rows   <- base::nrow(df.wide)
  port.val <- base::numeric(n.rows)
  port.ret <- base::numeric(n.rows)
  port.val[1L] <- 1

  for (i in 2L:n.rows) {
    # Grow each allocation by its return
    for (col in pair.cols) {
      allocs[col] <- allocs[col] * (1 + df.wide[[col]][i])
    }
    port.val[i] <- base::sum(allocs)
    port.ret[i] <- port.val[i] / port.val[i - 1L] - 1

    # Monthly rebalance: detect month boundary between day i and day i+1
    if (i < n.rows) {
      curr.m <- lubridate::floor_date(df.wide$date[i],        "month")
      next.m <- lubridate::floor_date(df.wide$date[i + 1L],   "month")
      if (next.m > curr.m) {
        allocs             <- base::rep(port.val[i] / n.p, n.p)
        base::names(allocs) <- pair.cols
      }
    }
  }

  df.wide %>%
    dplyr::mutate(
      ret   = port.ret,
      cumeq = base::cumprod(1 + ret)
    ) %>%
    dplyr::select(date, dplyr::all_of(pair.cols), ret, cumeq)
}
```

```{r}
#| include: false

# Reference pair: run with open prices for CC/CO/OC execution timing diagnostics
result.train <- strategy_pair(
  prices.wide = log.prices.wide.train,
  tickerA     = tickerA,
  tickerB     = tickerB,
  n.ma        = 60,
  z.entry     = 2,
  z.stop      = 3,
  open.wide   = log.open.wide.train
)
```

```{r}
#| include: false

# Full portfolio: all tradeable pairs, equal weight, monthly rebalancing
port.train <- portfolio_pairs(
  pairs.df    = pairs.tradeable,
  prices.wide = log.prices.wide.train,
  open.wide   = log.open.wide.train,
  n.ma        = 60,
  z.entry     = 2,
  z.stop      = 3
)
```

### Execution Timing

The strategy submits market-on-open (MOO) orders the day after a z-score signal fires. This avoids any look-ahead from the closing price that generated the signal. The return is computed using the appropriate convention per day:

-   **Entry day** - position opens at the next-day open; spread earns from open to close (OC)
-   **Holding days** - fully in position; spread earns the full close-to-close return (CC)
-   **Exit day** - position closes at the next-day open; spread earns only the overnight gap (CO)

```{r}
#| include: false

df.exec <- result.train %>%
  tidyr::drop_na(cumeq, cumeq.co, cumeq.oc) %>%
  dplyr::select(date, CC = cumeq, CO = cumeq.co, OC = cumeq.oc) %>%
  tidyr::pivot_longer(cols = c(CC, CO, OC), names_to = "Convention", values_to = "CumEQ")
```

```{r}
#| include: false

dplyr::tibble(
  Convention = c("Close → Close (CC)", "Close → Open (CO)", "Open → Close (OC)"),
  `Cum. Return` = c(
    base::round(dplyr::last(result.train$cumeq) - 1, 4),
    base::round(dplyr::last(result.train$cumeq.co) - 1, 4),
    base::round(dplyr::last(result.train$cumeq.oc) - 1, 4)
  )
) %>% gt::gt()
```

### Portfolio Equity - Default Parameters

```{r}
#| include: false

# Single-pair panels (hidden — kept for internal reference)
p1 <- result.train %>%
  tidyr::drop_na(z.score) %>%
  ggplot2::ggplot(ggplot2::aes(x = date, y = z.score)) +
  ggplot2::geom_line(col = "blue", linewidth = 0.3) +
  ggplot2::geom_hline(yintercept =  2, col = "red", linetype = "dashed") +
  ggplot2::geom_hline(yintercept = -2, col = "red", linetype = "dashed") +
  ggplot2::geom_hline(yintercept =  0, col = "grey20") +
  ggplot2::theme_minimal() +
  ggplot2::labs(title = paste0(tickerA, " / ", tickerB, " — Reference Pair"), x = "", y = "Z-Score")

p2 <- result.train %>%
  tidyr::drop_na(trade) %>%
  ggplot2::ggplot(ggplot2::aes(x = date, y = trade)) +
  ggplot2::geom_col(fill = "blue") +
  ggplot2::theme_minimal() +
  ggplot2::labs(x = "", y = "Trades")

p3 <- result.train %>%
  tidyr::drop_na(pos) %>%
  ggplot2::ggplot(ggplot2::aes(x = date, y = pos)) +
  ggplot2::geom_step(col = "blue") +
  ggplot2::geom_hline(yintercept = 0, col = "grey20") +
  ggplot2::theme_minimal() +
  ggplot2::labs(x = "", y = "Position")

p4 <- result.train %>%
  tidyr::drop_na(cumeq) %>%
  ggplot2::ggplot(ggplot2::aes(x = date, y = cumeq)) +
  ggplot2::geom_line(col = "blue") +
  ggplot2::geom_hline(yintercept = 1, col = "grey20") +
  ggplot2::theme_minimal() +
  ggplot2::labs(x = "", y = "CumEQ")
```

```{r}
#| echo: false
#| warning: false
#| message: false

# Portfolio equity curve — all tradeable pairs, equal weight, monthly rebalancing
port.train %>%
  tidyr::drop_na(cumeq) %>%
  ggplot2::ggplot(ggplot2::aes(x = date, y = cumeq)) +
  ggplot2::geom_line(col = "blue", linewidth = 0.8) +
  ggplot2::geom_hline(yintercept = 1, col = "grey20") +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title    = "Portfolio Equity — All Tradeable Pairs (Default Parameters)",
    subtitle = paste0(
      nrow(pairs.tradeable), " pairs | Equal weight | Monthly rebalancing",
      " | z.entry = 2, n.ma = 60 | Training: Jun 2000 – Dec 2024"
    ),
    x = "",
    y = "Cumulative Equity"
  )
```

**Interpretation - Portfolio Equity (Default Parameters):**

The portfolio combines all tradeable pairs into a single equity stream. Each pair contributes an independent mean-reversion signal, and monthly rebalancing continuously redirects capital toward pairs with room to revert and away from pairs that have already closed their spread. The default parameters (z.entry = 2, n.ma = 60) are conservative, the strategy is in the market only when the spread is dislocated by at least two standard deviations. The optimization section finds the parameter combination that maximizes the risk-adjusted return across the full portfolio.

------------------------------------------------------------------------

## Optimization

The strategy has two tunable parameters: **`z.entry`** (signal threshold) and **`n.ma`** (Kalman adaptation window). In this section, we will optimize the two parameters in order to get the best risk to reward ratio.

### Strategy Function and Parameter Grid

```{r}
#| echo: false
#| warning: false
#| message: false
#| cache: true

# Parameter grid (8.7.4.1: expand.grid approach) — fine mesh for 3D surface (8.8.9)
out <- base::expand.grid(
  z.entry = base::seq(0.50, 2.75, by = 0.25),
  n.ma    = base::seq(10L, 120L, by = 10L)
)

# Sweep: purrr::map2 — vectorize, not for-loops
# Each combination is evaluated on the full equal-weight portfolio
safe_trade_stats <- purrr::possibly(
  function(ze, nm) {
    portfolio_pairs(
      pairs.df    = pairs.tradeable,
      prices.wide = log.prices.wide.train,
      open.wide   = log.open.wide.train,
      n.ma        = nm,
      z.entry     = ze,
      z.stop      = 3
    ) %>%
      dplyr::select(date, ret) %>%
      RTL::tradeStats()
  },
  otherwise = NULL
)

res <- out %>%
  dplyr::mutate(
    stats = purrr::map2(z.entry, n.ma, safe_trade_stats)
  ) %>%
  dplyr::filter(!purrr::map_lgl(stats, is.null)) %>%
  tidyr::unnest_wider(stats)
```

### Optimization Visualization

```{r}
#| echo: false
#| warning: false
#| message: false

make_heatmap <- function(df, metric_col, title_str) {
  df %>%
    dplyr::select(z.entry, n.ma, value = dplyr::all_of(metric_col)) %>%
    dplyr::mutate(
      z.entry = base::factor(z.entry),
      n.ma    = base::factor(n.ma)
    ) %>%
    ggplot2::ggplot(ggplot2::aes(x = z.entry, y = n.ma, fill = value)) +
    ggplot2::geom_tile(col = "white") +
    ggplot2::scale_fill_gradient2(
      low      = "red",
      mid      = "white",
      high     = "blue",
      midpoint = 0
    ) +
    ggplot2::theme_minimal() +
    ggplot2::labs(
      title = title_str,
      x     = "Z-Score Entry Threshold",
      y     = "Kalman MA Window (n.ma)",
      fill  = ""
    )
}

p.sharpe  <- make_heatmap(res, "Sharpe",    "Sharpe Ratio")
p.cumret  <- make_heatmap(res, "CumReturn", "Cumulative Return")
p.ddmax   <- make_heatmap(res, "DD.Max",    "Max Drawdown")

(p.sharpe | p.cumret) / p.ddmax
```

**Interpretation - Optimization Heatmaps:**

The heatmaps show how the strategy's performance varies across the full parameter grid. Reading them together:

-   **Sharpe Ratio**: The highest Sharpe values cluster in the low `z.entry` and moderate-to-high `n.ma` region. Low z.entry means the strategy enters trades more frequently (smaller dislocation required). Higher n.ma means a smoother Kalman filter that adapts slowly — appropriate for a slow-reverting pair.
-   **Cumulative Return**: Mirrors the Sharpe pattern, more frequent trading (low z.entry) generates higher total return.
-   **Max Drawdown**: More aggressive parameters (lower z.entry) also produce deeper drawdowns. This is the fundamental risk-return tradeoff.

**Key insight:** The heatmaps do **not** show a single dominant corner, there is no "free lunch" parameter. Higher return always comes with higher drawdown. The 3D surfaces below make the tradeoff structure explicit.

### 3D Surface Visualization

Three interactive surfaces show the optimization landscape. We are looking for **a broad plateau** (flat-but-high region), not a sharp spike. A spike means the parameter is fragile; a plateau means nearby parameters produce similar performance, which is the hallmark of robustness.

```{r}
#| echo: false
#| warning: false
#| message: false

par1.vals <- base::sort(base::unique(res$z.entry))
par2.vals <- base::sort(base::unique(res$n.ma))

mat.sharpe <- res %>%
  dplyr::select(z.entry, n.ma, Sharpe) %>%
  tidyr::pivot_wider(names_from = n.ma, values_from = Sharpe) %>%
  dplyr::select(-z.entry) %>%
  base::as.matrix()

mat.cumret <- res %>%
  dplyr::select(z.entry, n.ma, CumReturn) %>%
  tidyr::pivot_wider(names_from = n.ma, values_from = CumReturn) %>%
  dplyr::select(-z.entry) %>%
  base::as.matrix()

mat.ddmax <- res %>%
  dplyr::select(z.entry, n.ma, DD.Max) %>%
  tidyr::pivot_wider(names_from = n.ma, values_from = DD.Max) %>%
  dplyr::select(-z.entry) %>%
  base::as.matrix()

plotly::plot_ly(x = ~par2.vals, y = ~par1.vals, z = ~mat.sharpe) %>%
  plotly::add_surface(colorscale = "Viridis") %>%
  plotly::layout(
    title = list(text = "Sharpe Ratio Surface"),
    scene = list(
      xaxis = list(title = "n.ma (Kalman Window)"),
      yaxis = list(title = "z.entry (Threshold)"),
      zaxis = list(title = "Sharpe")
    )
  )
```

```{r}
#| echo: false
#| warning: false
#| message: false

plotly::plot_ly(x = ~par2.vals, y = ~par1.vals, z = ~mat.cumret) %>%
  plotly::add_surface(colorscale = "Viridis") %>%
  plotly::layout(
    title = list(text = "Cumulative Return Surface"),
    scene = list(
      xaxis = list(title = "n.ma (Kalman Window)"),
      yaxis = list(title = "z.entry (Threshold)"),
      zaxis = list(title = "CumReturn")
    )
  )
```

```{r}
#| echo: false
#| warning: false
#| message: false

plotly::plot_ly(x = ~par2.vals, y = ~par1.vals, z = ~mat.ddmax) %>%
  plotly::add_surface(colorscale = "Viridis") %>%
  plotly::layout(
    title = list(text = "Maximum Drawdown Surface"),
    scene = list(
      xaxis = list(title = "n.ma (Kalman Window)"),
      yaxis = list(title = "z.entry (Threshold)"),
      zaxis = list(title = "DD.Max")
    )
  )
```

**Reading the 3D surfaces:**

-   **Sharpe surface**: Look for the broad, elevated region. If the peak is a sharp needle, those parameters are overfit. If the peak forms a wide mesa, those parameters are robust.
-   **Cumulative Return surface**: Shows total return across the grid. A rising slope along one axis means that axis dominates performance.
-   **Max Drawdown surface**: Higher values (closer to zero) are better. Deep valleys indicate parameter combinations where the strategy suffered severe losses.

The "sweet spot" is the parameter region that is simultaneously **high Sharpe, high CumReturn, and shallow drawdown**, which is a bit hard to formalize only looking at the above charts, that is why the next section will solve this using Z-score normalization.

### Z-Score Normalization

To compare metrics on the same scale, each performance dimension is standardized to a Z-score: (Z = \frac{x - \mu}{\sigma}). A positive Z-score means "better than average across the grid." The composite Z-score is the sum of all three: a parameter combination that is above-average on *every* dimension receives a high composite score.

```{r}
#| echo: false
#| warning: false
#| message: false

res.z <- res %>%
  dplyr::select(z.entry, n.ma, Sharpe, CumReturn, DD.Max) %>%
  tidyr::pivot_longer(
    cols      = c(Sharpe, CumReturn, DD.Max),
    names_to  = "variable",
    values_to = "value"
  ) %>%
  dplyr::group_by(variable) %>%
  dplyr::mutate(
    value.z = (value - base::mean(value, na.rm = TRUE)) /
              stats::sd(value, na.rm = TRUE)
  ) %>%
  dplyr::ungroup()

res.z %>%
  ggplot2::ggplot(ggplot2::aes(x = value.z)) +
  ggplot2::geom_histogram(
    col  = "black",
    fill = "blue",
    ggplot2::aes(y = ggplot2::after_stat(density)),
    bins = 25
  ) +
  ggplot2::facet_wrap(~ variable, scales = "free_y") +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title    = "8.8.9.1 Z-Score Distributions by Metric",
    subtitle = "Positive Z = above-average performance | Used for composite robustness scoring",
    x        = "Z-Score",
    y        = "Density"
  )
```

### Robust Parameter Selection - Plateau Method

**Method:** For each grid point, compute a composite Z-score (sum of Z-scored Sharpe + CumReturn + DD.Max). Then compute the average composite of its immediate neighbors (±1 grid step). The **robustness score** is the minimum of the two: `robustness = min(own composite, neighbor average)`. This max-min criterion penalizes isolated spikes and rewards broad plateaus.

```{r}
#| echo: false
#| warning: false
#| message: false

composite <- res.z %>%
  dplyr::group_by(z.entry, n.ma) %>%
  dplyr::summarise(composite.z = base::sum(value.z), .groups = "drop")

z.step  <- 0.25
ma.step <- 10L

composite <- composite %>%
  dplyr::mutate(
    neighbor.avg = purrr::map2_dbl(z.entry, n.ma, function(ze, nm) {
      neighbors <- composite %>%
        dplyr::filter(
          base::abs(z.entry - ze) <= z.step,
          base::abs(n.ma - nm)    <= ma.step,
          !(z.entry == ze & n.ma == nm)
        )
      if (base::nrow(neighbors) == 0L) return(NA_real_)
      base::mean(neighbors$composite.z, na.rm = TRUE)
    }),
    robustness = base::pmin(composite.z, neighbor.avg)
  )

robust.pick <- composite %>%
  dplyr::filter(!base::is.na(robustness)) %>%
  dplyr::arrange(dplyr::desc(robustness)) %>%
  dplyr::slice(1)
```

```{r}
#| echo: false
#| warning: false
#| message: false

# Composite surface — the plateau is visible as a broad elevated region
mat.composite <- composite %>%
  dplyr::select(z.entry, n.ma, composite.z) %>%
  tidyr::pivot_wider(names_from = n.ma, values_from = composite.z) %>%
  dplyr::select(-z.entry) %>%
  base::as.matrix()

plotly::plot_ly(x = ~par2.vals, y = ~par1.vals, z = ~mat.composite) %>%
  plotly::add_surface(colorscale = "Viridis") %>%
  plotly::layout(
    title = list(text = "Composite Z-Score Surface (Sharpe + CumReturn + DD.Max)"),
    scene = list(
      xaxis = list(title = "n.ma"),
      yaxis = list(title = "z.entry"),
      zaxis = list(title = "Composite Z")
    )
  )
```

```{r}
#| echo: false
#| warning: false
#| message: false

# Highlight the robust pick and top-5 plateau points
composite %>%
  dplyr::filter(!base::is.na(robustness)) %>%
  dplyr::arrange(dplyr::desc(robustness)) %>%
  dplyr::slice(1:5) %>%
  dplyr::left_join(
    res %>% dplyr::select(z.entry, n.ma, Sharpe, CumReturn, DD.Max),
    by = c("z.entry", "n.ma")
  ) %>%
  dplyr::select(z.entry, n.ma, Sharpe, CumReturn, DD.Max, composite.z, robustness) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "8.8.9.2 Top-5 Robust Parameter Combinations",
    subtitle = "Ranked by min(own composite, neighbor avg) — plateau criterion"
  ) %>%
  gt::fmt_number(
    columns  = c(Sharpe, CumReturn, DD.Max, composite.z, robustness),
    decimals = 3
  ) %>%
  gt::tab_style(
    style     = gt::cell_fill(color = "#E8F5E9"),
    locations = gt::cells_body(rows = 1)
  ) %>%
  gt::tab_footnote(
    footnote  = "Green row = selected parameters. Plateau method guards against overfitting: isolated spikes are penalized.",
    locations = gt::cells_title(groups = "subtitle")
  )
```

**Interpretation - Robust Selection:**

The composite Z-score surface combines Sharpe, cumulative return, and drawdown into a single "desirability" metric. The plateau method selects from the broad elevated region rather than the absolute peak. **This is critical for out-of-sample stability: if the surface is jagged and we pick the tallest spike, we are likely fitting noise. If we pick from a mesa, nearby parameters would have produced similar results a.k.a the strategy is not fragile to small perturbations in parameter choice.**

### Risk Appetite - Select Optimal Parameters

Following the plateau robustness analysis above: the selected parameters are the most robust combination from the composite Z-score surface. As a secondary check, they must also satisfy a risk appetite floor (Sharpe \> 0 and Max Drawdown \> −0.50).

```{r}
#| echo: false
#| warning: false
#| message: false

# Use robust plateau pick from 8.8.9.2
optimal <- res %>%
  dplyr::filter(
    z.entry == robust.pick$z.entry,
    n.ma    == robust.pick$n.ma
  ) %>%
  dplyr::slice(1)

optimal %>%
  dplyr::select(z.entry, n.ma, CumReturn, Ret.Ann, SD.Ann, Sharpe, Omega, `%.Win`, DD.Max) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "8.9 Selected Parameters — Risk Appetite Filter",
    subtitle = "Selected via plateau robustness method (8.8.9.2) — flat-but-high across all 3 metrics"
  ) %>%
  gt::fmt_number(
    columns  = c(CumReturn, Ret.Ann, SD.Ann, Sharpe, Omega, `%.Win`, DD.Max),
    decimals = 4
  )
```

```{r}
#| include: false

# Re-run full portfolio with selected optimal parameters
result.opt <- portfolio_pairs(
  pairs.df    = pairs.tradeable,
  prices.wide = log.prices.wide.train,
  open.wide   = log.open.wide.train,
  n.ma        = optimal$n.ma,
  z.entry     = optimal$z.entry,
  z.stop      = 3
)
```

## Risk / Reward Measures

### Sharpe Ratio

\[SR = \frac{r - r_f}{\sigma}\]

The Sharpe ratio is a parametric measure of risk-reward: excess return per unit of volatility. When assuming `r_f = 0`, it measures actual returns per unit of risk.

```{r}
#| echo: false
#| warning: false
#| message: false

ret.xts <- xts::xts(
  result.opt$ret,
  order.by = result.opt$date
)

sr <- PerformanceAnalytics::SharpeRatio(ret.xts, Rf = 0, FUN = "StdDev")

dplyr::tibble(
  Metric = base::rownames(sr),
  Value  = base::as.numeric(sr)
) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "8.8.1 Sharpe Ratio — Optimal Parameters (Training)",
    subtitle = "Multi-pair equal-weight portfolio | Rf = 0"
  ) %>%
  gt::fmt_number(columns = Value, decimals = 4)
```

### Omega Ratio

The Omega ratio measures the ratio of gains above a threshold to losses below it. `Omega > 1` means more upside than downside. Omega also considers the effect of fat tails, whereas Sharpe just assumes that returns are normally distributed.

```{r}
#| echo: false
#| warning: false
#| message: false

omega.val <- PerformanceAnalytics::Omega(ret.xts, L = 0)
omegarsr  <- PerformanceAnalytics::OmegaSharpeRatio(ret.xts)

# Return distribution
ret.vec <- base::as.numeric(ret.xts[ret.xts != 0])

p.omega <- dplyr::tibble(ret = ret.vec) %>%
  ggplot2::ggplot(ggplot2::aes(x = ret)) +
  ggplot2::geom_histogram(
    col  = "black",
    fill = "blue",
    ggplot2::aes(y = ggplot2::after_stat(density)),
    bins = 40
  ) +
  ggplot2::stat_function(
    fun  = stats::dnorm,
    args = list(
      mean = base::mean(ret.vec, na.rm = TRUE),
      sd   = stats::sd(ret.vec, na.rm = TRUE)
    ),
    col  = "red",
    linewidth = 1
  ) +
  ggplot2::geom_vline(xintercept = 0, col = "black") +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title    = paste0(
      "8.8.2 Sharpe vs Omega | Omega = ", base::round(base::as.numeric(omega.val), 3),
      " | Omega-Sharpe = ", base::round(base::as.numeric(omegarsr), 3)
    ),
    subtitle = "Multi-pair portfolio daily returns | Blue bars = empirical | Red curve = normal",
    x        = "Daily Return",
    y        = "Density"
  )

p.omega
```

Comparing Omega to Sharpe, we can see that our strategy has more return centered around 0, and has a fatter tails. This is why the two ratios diverge.

### Full Performance

```{r}
#| echo: false
#| warning: false
#| message: false

stats.train <- result.opt %>%
  dplyr::select(date, ret) %>%
  RTL::tradeStats()

dplyr::as_tibble(stats.train) %>%
  tidyr::pivot_longer(
    cols      = dplyr::everything(),
    names_to  = "Metric",
    values_to = "Value"
  ) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "8.4.5 Performance Assessment — Training Period",
    subtitle = paste0(
      "Multi-pair equal-weight portfolio | Monthly rebalancing",
      " | z.entry = ", optimal$z.entry,
      " | n.ma = ", optimal$n.ma
    )
  ) %>%
  gt::fmt_number(columns = Value, decimals = 4)
```

### Drawdown Analysis

Drawdowns measure the percentage loss since the prior peak. Max Drawdown Length indicates how long the strategy has been underwater.

```{r}
#| echo: false
#| warning: false
#| message: false

# Drawdown table (top 10 worst)
dd.table <- PerformanceAnalytics::table.Drawdowns(ret.xts, top = 10)

dd.table %>%
  dplyr::as_tibble() %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "8.8.6 Top 10 Drawdowns — Training Period",
    subtitle = "Multi-pair equal-weight portfolio | Optimal parameters"
  ) %>%
  gt::fmt_number(columns = Depth, decimals = 4)
```

```{r}
#| echo: false
#| warning: false
#| message: false

# Drawdown chart (ggplot2 equivalent of PerformanceAnalytics::chart.Drawdown)
cumeq.vec <- result.opt$cumeq
running.max <- base::cummax(cumeq.vec)
drawdown    <- cumeq.vec / running.max - 1

dplyr::tibble(
  date     = result.opt$date,
  drawdown = drawdown
) %>%
  tidyr::drop_na() %>%
  ggplot2::ggplot(ggplot2::aes(x = date, y = drawdown)) +
  ggplot2::geom_area(fill = "red", alpha = 0.4) +
  ggplot2::geom_line(col = "red") +
  ggplot2::geom_hline(yintercept = 0, col = "black") +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title    = "8.8.6–8.8.7 Drawdown — Multi-pair Portfolio",
    subtitle = "Percentage loss from prior equity peak | Training period",
    x        = "",
    y        = "Drawdown"
  )
```

### Correlation to Underlying (SPY Benchmark)

When correlation ρ = 1, the strategy returns equal long buy-and-hold. When ρ = −1, it equals short buy-and-hold. A market-neutral pairs strategy should have ρ near zero, confirming it does not simply replicate a directional SPY position.

```{r}
#| echo: false
#| warning: false
#| message: false

# SPY returns for benchmark comparison
spy.ret <- etf.ret.train %>%
  dplyr::filter(symbol == "SPY") %>%
  dplyr::select(date, ret.spy = ret)

strat.ret <- result.opt %>%
  dplyr::select(date, ret) %>%
  tidyr::drop_na()

df.corr <- strat.ret %>%
  dplyr::inner_join(spy.ret, by = "date")

cor.val <- stats::cor(df.corr$ret, df.corr$ret.spy, use = "complete.obs")

p.corr <- df.corr %>%
  ggplot2::ggplot(ggplot2::aes(x = ret.spy, y = ret)) +
  ggplot2::geom_point(col = "grey20", alpha = 0.3, size = 0.8) +
  ggplot2::geom_smooth(method = "lm", col = "red", se = TRUE) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title    = paste0(
      "8.8.8 Portfolio vs SPY Correlation | ρ = ",
      base::round(cor.val, 3)
    ),
    subtitle = "Near-zero correlation confirms market neutrality | Red = linear fit",
    x        = "SPY Daily Return",
    y        = "Multi-pair Portfolio Return"
  )

p.corr
```

------------------------------------------------------------------------

## Out-of-Sample Evaluation

Parameters selected in from the Optimization section are applied to the **test period (Jan 2025 – Mar 2026)** without any refitting. All model parameters are fit exclusively on training data, i.e. the test period is never touched until final evaluation.

```{r}
#| echo: false
#| warning: false
#| message: false

result.test <- portfolio_pairs(
  pairs.df    = pairs.tradeable,
  prices.wide = log.prices.wide.test,
  open.wide   = log.open.wide.test,
  n.ma        = optimal$n.ma,
  z.entry     = optimal$z.entry,
  z.stop      = 3
)

stats.test <- result.test %>%
  dplyr::select(date, ret) %>%
  RTL::tradeStats()
```

```{r}
#| echo: false
#| warning: false
#| message: false

# Train vs Test comparison table
dplyr::bind_rows(
  dplyr::as_tibble(stats.train) %>%
    tidyr::pivot_longer(dplyr::everything(), names_to = "Metric", values_to = "Train"),
  dplyr::as_tibble(stats.test) %>%
    tidyr::pivot_longer(dplyr::everything(), names_to = "Metric", values_to = "Test")
) %>%
  dplyr::group_by(Metric) %>%
  dplyr::summarise(
    Train = dplyr::first(Train),
    Test  = dplyr::last(Test),
    .groups = "drop"
  ) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "8.10 Train vs Out-of-Sample Performance",
    subtitle = paste0(
      "Multi-pair equal-weight portfolio | Monthly rebalancing",
      " | z.entry = ", optimal$z.entry,
      " | n.ma = ", optimal$n.ma
    )
  ) %>%
  gt::fmt_number(columns = c(Train, Test), decimals = 4)
```

```{r}
#| echo: false
#| warning: false
#| message: false

# Out-of-sample portfolio equity curve
result.test %>%
  tidyr::drop_na(cumeq) %>%
  ggplot2::ggplot(ggplot2::aes(x = date, y = cumeq)) +
  ggplot2::geom_line(col = "blue", linewidth = 0.8) +
  ggplot2::geom_hline(yintercept = 1, col = "grey20") +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title    = "8.10 Out-of-Sample Portfolio Equity",
    subtitle = paste0(
      "Jan 2025 – Mar 2026 | Parameters selected on training data only",
      " | z.entry = ", optimal$z.entry, " | n.ma = ", optimal$n.ma
    ),
    x = "",
    y = "Cumulative Equity"
  )
```

### Interpretation - Out-of-Sample Results

**What the numbers actually say:**

The test period produced a Sharpe of 0.2145 and an annualized return of 1.00%. To understand what that means, consider the context: the US risk-free rate in 2025 sat between 4.25% and 4.50%. A strategy that earns 1% annualized is not just underperforming equity benchmarks, it is losing to cash. If Sharpe is recalculated using a realistic Rf of 0.045, the risk-adjusted return flips negative. The Sharpe of 0.2145 only exists because the formula was applied with Rf = 0.

The win rate (0.4619) and time-in-market (0.6355) held close to training levels, which confirms the signal logic is mechanically consistent. The pipeline is not broken. The problem is that the alpha it generates, even when the mechanics work as designed, is too thin to matter. The strategy spent 63% of the test period in the market and earned less than sitting in a money market fund.

The maximum drawdown of 3.44% looks clean relative to training's 19.52%, but the comparison is partly misleading. Fifteen months is a short enough window that deep drawdowns rarely accumulate. The shallow drawdown says less about the strategy's risk management than about the brevity of the test period.

**Why the return is this low:**

The sector ETF universe is small, liquid, and heavily covered by institutional capital. Any mean-reversion relationship in it has been arbitraged significantly over the past two decades. The pairs identified by this pipeline are real and cointegration confirms that, but the speed and magnitude of reversion has compressed to the point where gross returns barely cover the theoretical opportunity. Transaction costs, which were not modeled, would likely push net return into negative territory.

The 2025 macro environment added another layer. Tariff shocks and policy-driven sector rotation created regime breaks that were outside anything the optimizer trained on. Sector spreads that might normally revert over weeks widened and stayed wide for months, reducing the number of complete round-trip cycles during the test window.

**What this means for the strategy:**

The 15-month result is not a failure of execution. It is an honest signal that this strategy, applied to large-cap US sector ETFs under current market conditions, does not generate economically meaningful alpha. The structural mechanics work. The returns do not justify the complexity. The benchmark comparison in the next section makes the gap concrete.

------------------------------------------------------------------------

## Benchmark Comparison

A market-neutral strategy should not be evaluated in isolation. The relevant question is not whether the strategy made money, but whether it made enough money to justify holding it instead of something simpler. This section compares the out-of-sample strategy return directly against two benchmarks over the same January 2025 to March 2026 window: SPY as the broad market proxy, and an equal-weight portfolio of all nine sector ETFs used in the clustering step (XLB, XLE, XLF, XLI, XLK, XLP, XLV, XLU, XLY) with monthly rebalancing.

The equal-weight sector index is a deliberate choice. It uses the exact same building blocks as the strategy with zero modeling overhead. If that naive alternative beats the strategy, it is a direct comment on whether the complexity added any value.

```{r}
#| echo: false
#| warning: false
#| message: false

# Equal-weight sector index -- same 9 ETFs, monthly rebalancing
tickers.ew <- c("XLB", "XLE", "XLF", "XLI", "XLK", "XLP", "XLV", "XLU", "XLY")

ew.daily <- etf.ret.test %>%
  dplyr::filter(symbol %in% tickers.ew) %>%
  dplyr::group_by(date) %>%
  dplyr::summarise(ret = base::mean(ret, na.rm = TRUE), .groups = "drop") %>%
  dplyr::mutate(cumeq = base::cumprod(1 + ret))

# SPY over test period
spy.test <- etf.ret.test %>%
  dplyr::filter(symbol == "SPY") %>%
  dplyr::select(date, ret) %>%
  dplyr::mutate(cumeq = base::cumprod(1 + ret))

# Align all three series on shared dates
df.bench <- result.test %>%
  dplyr::select(date, cumeq.strat = cumeq) %>%
  dplyr::inner_join(
    spy.test %>% dplyr::select(date, cumeq.spy = cumeq),
    by = "date"
  ) %>%
  dplyr::inner_join(
    ew.daily %>% dplyr::select(date, cumeq.ew = cumeq),
    by = "date"
  ) %>%
  tidyr::drop_na()
```

```{r}
#| echo: false
#| warning: false
#| message: false

# Equity curve overlay
df.bench %>%
  tidyr::pivot_longer(
    cols      = c(cumeq.strat, cumeq.spy, cumeq.ew),
    names_to  = "series",
    values_to = "cumeq"
  ) %>%
  dplyr::mutate(
    series = dplyr::case_when(
      series == "cumeq.strat" ~ "Strategy",
      series == "cumeq.spy"   ~ "SPY",
      series == "cumeq.ew"    ~ "Equal-Weight Sectors"
    )
  ) %>%
  ggplot2::ggplot(ggplot2::aes(x = date, y = cumeq, col = series)) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::scale_color_manual(
    values = c("Strategy" = "blue", "SPY" = "red", "Equal-Weight Sectors" = "green")
  ) +
  ggplot2::geom_hline(yintercept = 1, col = "grey20", linetype = "dashed") +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title    = "Strategy vs Benchmarks — Out-of-Sample",
    subtitle = "Jan 2025 – Mar 2026 | Strategy (blue) vs SPY (red) vs Equal-Weight Sectors (green)",
    x        = "",
    y        = "Cumulative Equity",
    col      = ""
  )
```

```{r}
#| echo: false
#| warning: false
#| message: false

# Performance metrics for each benchmark
stats.spy.bench <- spy.test %>%
  dplyr::select(date, ret) %>%
  RTL::tradeStats()

stats.ew.bench <- ew.daily %>%
  dplyr::select(date, ret) %>%
  RTL::tradeStats()

# Side-by-side comparison table
dplyr::tibble(
  Metric         = base::names(dplyr::as_tibble(stats.test)),
  Strategy       = base::unlist(dplyr::as_tibble(stats.test)),
  SPY            = base::unlist(dplyr::as_tibble(stats.spy.bench)),
  Equal.Weight   = base::unlist(dplyr::as_tibble(stats.ew.bench))
) %>%
  dplyr::filter(Metric %in% c("CumReturn", "Ret.Ann", "SD.Ann", "Sharpe", "Omega", "DD.Max", "%.Win")) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Performance Comparison — Out-of-Sample",
    subtitle = "Jan 2025 – Mar 2026 | Rf = 0 for all Sharpe calculations"
  ) %>%
  gt::fmt_number(columns = c(Strategy, SPY, Equal.Weight), decimals = 4) %>%
  gt::tab_footnote(
    footnote  = "Equal-weight index holds all 9 clustering ETFs at equal weight with monthly rebalancing. No signal, no filter.",
    locations = gt::cells_column_labels(columns = Equal.Weight)
  )
```

The equity curves and metrics table together answer the core question. The strategy's 1% annualized return sits well below both benchmarks. The equal-weight sector index, which requires no model, no optimization, and no Kalman filter, delivers materially higher returns over the same window. SPY compounds that gap further. Market neutrality is a real property of the strategy but in a bull market regime it means systematically forgoing the beta that was the primary driver of returns in 2025 and early 2026. The pairs strategy earns its low drawdown by construction. The cost is that it earns almost nothing else.

------------------------------------------------------------------------

## Literature Flow — Which Paper Supports Which Step

Each step in the strategy pipeline is grounded in academic literature. The diagram below traces the intellectual lineage: which paper motivated which modeling choice, and where the class notes provide direct instruction versus where external research was required.

```{mermaid}
graph LR
    subgraph PAPERS["<b>Academic Sources</b>"]
        P1["<b>Cartea, Cucuringu & Jin (2023)</b><br/>SSRN 4560455<br/>Correlation-matrix clustering<br/>→ market-neutral portfolios"]
        P2["<b>Yang, Huang & Chen (2023)</b><br/>Hierarchical clustering +<br/>Kalman filter blueprint"]
        P3["<b>Simonson (2018)</b><br/>OU process for ETF spreads<br/>University of Washington"]
        P4["<b>Avellaneda & Lee (2008)</b><br/>SSRN 1153505<br/>Statistical arbitrage<br/>Sharpe 1.44–1.51"]
        P5["<b>Engle & Granger (1987)</b><br/>Nobel Prize — cointegration<br/>ADF residual test"]
        P6["<b>Kalman (1960)</b><br/>State-space filtering<br/>Optimal linear estimator"]
    end

    subgraph CLASS["<b>FIN 452 Class Notes</b>"]
        C1["Module 2<br/>Trading Process"]
        C2["Module 3<br/>Sector Rotation"]
        C3["Module 8<br/>Quantitative Trading"]
    end

    subgraph STEPS["<b>Strategy Steps</b>"]
        S1["Step 1: Clustering"]
        S2["Step 2: Cointegration"]
        S3["Step 3: OU Half-Life"]
        S4["Step 4: Kalman Filter"]
        S5["8.4 Strategy Function"]
        S6["8.7 Optimization"]
        S7["8.8 Risk / Reward"]
    end

    P1 --> S1
    P2 --> S1
    P2 --> S2
    P2 --> S4
    P3 --> S3
    P4 --> S4
    P5 --> S2
    P6 --> S4
    C1 --> S5
    C2 --> S1
    C3 --> S5
    C3 --> S6
    C3 --> S7
```

```{r}
#| echo: false
#| warning: false
#| message: false

dplyr::tibble(
  `Strategy Step` = c(
    "Step 1 — Clustering",
    "Step 1 — Clustering",
    "Step 2 — Cointegration",
    "Step 2 — Cointegration",
    "Step 3 — OU Half-Life",
    "Step 4 — Kalman Filter",
    "Step 4 — Kalman Filter",
    "Step 4 — Kalman Filter"
  ),
  Source = c(
    "Cartea, Cucuringu & Jin (2023)",
    "Yang, Huang & Chen (2023)",
    "Engle & Granger (1987)",
    "Yang, Huang & Chen (2023)",
    "Simonson (2018)",
    "Kalman (1960)",
    "Avellaneda & Lee (2008)",
    "Yang, Huang & Chen (2023)"
  ),
  `Key Contribution` = c(
    "Clustering on correlation matrix → mean-reverting portfolios; Sharpe > 1 demonstrated on equities",
    "Hierarchical clustering as Layer 1 before cointegration + Kalman; direct blueprint for this strategy",
    "Cointegration theory and the ADF test for residual stationarity — Nobel Prize foundational work",
    "Cointegration as the statistical gate between clustering and Kalman — 'the two steps are complements'",
    "OU process calibration for ETF pairs; half-life as holding-period filter; sector-ETF focus",
    "Original state-space filtering paper; optimal linear estimator for time-varying hedge ratio",
    "Full statistical arbitrage framework on U.S. equities; Sharpe 1.44–1.51; historical benchmark",
    "End-to-end workflow: clustering → cointegration → Kalman → trading signal in a single paper"
  )
) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Literature Map — Academic and Class Sources by Strategy Step",
    subtitle = "Each row traces how a specific paper or class note motivates a modeling choice"
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_body(columns = `Strategy Step`)
  )
```

**How the papers connect:** Yang, Huang & Chen (2023) is the closest single-paper blueprint — it uses hierarchical clustering as a pre-filter, cointegration as a validation gate, and a Kalman filter for dynamic hedge ratios, exactly as implemented here. Cartea et al. (2023) provides the theoretical justification for *why* correlation-matrix clustering produces mean-reverting portfolios. Simonson (2018) fills the OU half-life gap not covered in class notes. Avellaneda & Lee (2008) is the historical anchor — it demonstrates that statistical arbitrage on U.S. equities can achieve Sharpe \> 1.4, setting the performance benchmark. Engle & Granger (1987) and Kalman (1960) are foundational — they provide the econometric and filtering tools that make the strategy mathematically rigorous.

------------------------------------------------------------------------

## Weaknesses and Limitations

No backtested strategy should be presented without an honest assessment of what could go wrong. The weaknesses below are organized from most to least severe.

```{r}
#| echo: false
#| warning: false
#| message: false

dplyr::tibble(
  `#` = 1:9,
  Weakness = c(
    "Slow mean-reversion",
    "No transaction costs",
    "Weak cluster separation",
    "Overfitting risk in optimization",
    "Small, homogeneous universe",
    "Fixed train/test split",
    "No regime detection",
    "Look-ahead in data construction",
    "Short out-of-sample window"
  ),
  Severity = c(
    "High", "High", "Medium", "Medium",
    "Medium", "Low", "Low", "Low", "Low"
  ),
  Description = c(
    "Sector ETF pairs revert over months, producing very few round trips per year. The 15-month test window likely contains only 2 to 4 complete cycles, which is too few for statistical confidence.",
    "Bid/ask spreads, commissions, and slippage are not modeled. The strategy earns approximately 1% annualized gross out of sample. At that return level, even minimal ETF bid/ask spreads applied across 63% time-in-market across multiple pairs would likely eliminate most or all of the net return. This is not a theoretical concern.",
    "Silhouette score of 0.16 is well below the 0.5 threshold. All ETFs share heavy SPY exposure, making clean separation impossible. Clustering is organizational; cointegration is the true filter.",
    "A 120-point grid optimized on 24 years of data carries overfitting risk. The plateau method mitigates but does not eliminate the chance that selected parameters fail in future regimes.",
    "Ten U.S. large-cap sectors from one index are structurally correlated. A broader universe of international ETFs, factor ETFs, or commodities would yield more independent pair candidates.",
    "A single train/test split does not test across multiple regimes. Rolling or expanding window walk-forward validation would be more rigorous but was outside the scope of this assignment.",
    "The Kalman filter adapts the hedge ratio but cannot detect when cointegration has structurally broken. A regime-switching model such as Markov or CUSUM could flag breakdown periods.",
    "The split date is chosen a priori. While no test-period data leaks into model fitting, the split date choice itself could be optimized, which is a subtle form of look-ahead.",
    "Fifteen months contains roughly 3 half-life cycles at best. Any conclusion about out-of-sample performance, positive or negative, carries very wide confidence intervals."
  )
) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Weaknesses and Limitations",
    subtitle = "Honest assessment of what could go wrong and what was not modeled"
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(color = "red", weight = "bold"),
    locations = gt::cells_body(columns = Severity, rows = Severity == "High")
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(color = "orange", weight = "bold"),
    locations = gt::cells_body(columns = Severity, rows = Severity == "Medium")
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(color = "blue"),
    locations = gt::cells_body(columns = Severity, rows = Severity == "Low")
  )
```

**Key takeaway:** The remaining highest-severity weakness is slow reversion, which is intrinsic to the sector ETF universe. The practical implication is that this strategy requires a multi-year evaluation window to assess properly, and the 15-month test period in this assignment is insufficient for statistical significance.

------------------------------------------------------------------------

## Lessons Learned

These are a few lesson that I've learned from building this project.

-   Backtesting is optimistic by design. A 24-year training window is long enough to pick up real patterns, but also long enough for the optimizer to stumble onto patterns that are not actually there. The out-of-sample results made that gap very clear.

-   Every paper I referenced had in-sample Sharpe ratios above 1.0. The hard part was never building the pipeline. It was running it on data that had no idea I spent months optimizing on something else.

-   A more complex model needs better results to justify itself, not better methodology. This strategy has four filtering steps, a state-space model, a multi-objective optimizer, and a plateau robustness criterion. All of it has legitimate academic support.

-   But none of that matters if the strategy barely beats a savings account. Sophistication has to actually produce alpha that clears the bar set by simpler alternatives. This one did not clear that bar.

-   Knowing that the methodology is sound is not the same as knowing the results are good. That distinction matters a lot in practice. Someone who cannot be honest about their own output is a problem regardless of how sophisticated the model is.

-   The test window included tariff shocks, rate pivots, and some unusual sector rotation. A different 15-month stretch on the same code might have returned a Sharpe of 0.8. That is not an excuse, it is just a real constraint.

-   One test window does not prove anything either way. The right call is to present the numbers honestly and not pretend one draw from a wide distribution settles the question.

------------------------------------------------------------------------

## Conclusion

This project implements the full quantitative pairs trading workflow, starting from the hypothesis that sector ETFs share systematic risk factors that create exploitable mean-reverting spreads, then filtering the universe through a four-step pipeline (K-Means Clustering, Engle-Granger Cointegration, Ornstein-Uhlenbeck Half-Life, and Kalman-Filtered Dynamic Hedging) to identify tradeable pairs with statistically validated mean-reversion properties.

Out of sample, the strategy delivered a Sharpe ratio of 0.21 and an annualized return of 1.0% with a maximum drawdown of 3.4% over the January 2025 to March 2026 test period. Against a risk-free rate of approximately 4.5%, the risk-adjusted return on a real-cost basis is negative. SPY and a simple equal-weight basket of the same sector ETFs both outperformed significantly over the same window. The pipeline is methodologically sound, but the returns it produces in the large-cap US sector ETF universe do not justify the complexity relative to simpler alternatives. The structural mechanics work. The alpha does not clear the bar.

------------------------------------------------------------------------

## Things to Improve on

I intend to continue this project in the summer as this is super interesting!!

-   Asset classes: Instead of just operating on ETF that has a lot eyes watching it, I would love to try this strategy on less inefficient commodities, as I think there are more justification to pairs trade than an arbitrary regression! Even if I want to stick with Equities, I will try to test it on individual stocks

-   Clustering: Instead of clustering based only on Log Returns, I would love to find a way to make a composite factor through PCA that integrates idiosyncratic and sentiment
