---
title: "FIN 452 — Quantitative Trading Strategy"
subtitle: "Clustering + Kalman Filter | S&P 500 Select Sector ETFs"
author: "Ken Sucento"
date: "April 7, 2026"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    theme: flatly
    number-sections: true
---

```{r}
#| echo: false
#| warning: false
#| message: false
library(tidyverse)
library(tidyquant)
library(tidymodels)
library(rlang)
library(purrr)
library(lubridate)
library(plotly)
library(patchwork)
library(gt)
library(cluster)
library(factoextra)
library(tseries)
library(KFAS)
library(broom)
library(slider)
library(TTR)
library(PerformanceAnalytics)
library(RTL)
```

## Mental Model — How the Strategy Came to Life

The strategy answers one question: *"Can two sector ETFs that historically move together be traded profitably when their spread temporarily diverges?"* The answer requires four sequential filters — each one narrows the universe from many candidates to one actionable pair and a concrete trading rule. The diagram below shows the full lifecycle from hypothesis to evaluation.

```{mermaid}
graph TD
    subgraph HYPOTHESIS["<b>Hypothesis</b>"]
        H1["Sector ETFs share systematic risk<br/>→ their spreads should mean-revert"]
    end

    subgraph DATA["<b>Data Layer</b>"]
        D1["tidyquant::tq_get<br/>12 SPDR ETFs | Jun 2000 – Mar 2026<br/>Adjusted Close + Open Prices"]
        D2["Train / Test Split<br/>Train: Jun 2000 – Dec 2024<br/>Test: Jan 2025 – Mar 2026<br/>No information leakage"]
        D3["Log Returns + Log Prices<br/>Log returns for clustering<br/>Log prices for cointegration & Kalman"]
    end

    subgraph FILTER["<b>4-Step Filter Pipeline (Train Only)</b>"]
        F1["<b>Step 1: Clustering</b><br/>K-Means k = 3 on return correlations<br/>→ Defensive / Value-Cyclical / Growth<br/>Silhouette validation + stability check"]
        F2["<b>Step 2: Cointegration</b><br/>Engle-Granger ADF test<br/>All within-cluster pairs<br/>→ Keep p-value &lt; 0.05"]
        F3["<b>Step 3: OU Half-Life</b><br/>Regress Δspread ~ spread_lag<br/>→ Keep HL ≤ 252 days<br/>All tradeable pairs selected"]
        F4["<b>Step 4: Kalman Filter + Portfolio</b><br/>Dynamic hedge ratio β_t per pair<br/>Equal-weight portfolio, monthly rebalancing<br/>→ Portfolio daily return stream"]
    end

    subgraph SIGNAL["<b>Signal & Execution (8.4)</b>"]
        S1["Z-Score Signal<br/>Rolling 20-day mean and SD"]
        S2{"Trade Decision"}
        S3["Long Spread: Z &lt; −z.entry"]
        S4["Short Spread: Z &gt; +z.entry"]
        S5["Exit: return to flat"]
        S6["Stop-Loss: |Z| &gt; z.stop"]
        S7["8.4.1 Execution Timing<br/>CC / CO / OC decomposition"]
    end

    subgraph EVALUATE["<b>Evaluation (8.6 – 8.10)</b>"]
        E1["8.7 Optimization<br/>Grid search: z.entry × n.ma<br/>RTL::tradeStats per combo"]
        E1b["8.8.9 3D Surface Plots<br/>Sharpe · CumReturn · DD.Max<br/>Interactive plotly surfaces"]
        E1c["8.8.9.1 Z-Score Normalization<br/>Standardize all metrics<br/>Composite desirability score"]
        E1d["8.8.9.2 Plateau Selection<br/>max-min(own, neighbor avg)<br/>Anti-overfit robustness"]
        E2["8.9 Risk Appetite Filter<br/>Robust params from plateau"]
        E3["8.8 Risk / Reward<br/>Sharpe · Omega · Drawdown<br/>Correlation to SPY benchmark"]
        E4["8.10 Out-of-Sample<br/>Apply optimal params to test set<br/>No refitting — walk-forward"]
    end

    subgraph REFLECT["<b>Reflection</b>"]
        R1["Literature Flow<br/>Which paper → which step"]
        R2["Weaknesses<br/>10 limitations ranked by severity"]
    end

    H1 --> D1 --> D2 --> D3
    D3 --> F1 --> F2 --> F3 --> F4
    F4 --> S1 --> S2
    S2 --> S3 & S4 & S5 & S6
    S3 & S4 & S5 & S6 --> S7
    S7 --> E1 --> E1b --> E1c --> E1d --> E2 --> E3 --> E4
    E4 --> R1 & R2
```

**Reading the map:** Each box is a stage. Each arrow is a dependency — a later stage cannot run without the earlier stage's output. The filter pipeline progressively narrows: 10 ETFs → 22 candidate pairs → cointegrated survivors → tradeable pairs. All tradeable pairs are then run simultaneously as an equal-weight portfolio with monthly rebalancing, and that portfolio is stress-tested across parameter combinations and unseen data.

## Data

### ETF Universe

The strategy universe consists of eleven S&P 500 Select Sector SPDR ETFs plus the benchmark (SPY). Two ETFs have limited history and are **excluded from the clustering model**:

-   **XLC** (Communication Services) — available from June 2018 only; only \~6.5 years of training data
-   **XLRE** (Real Estate) — available from October 2015 only; only \~9 years of training data

Both are retained in the full data pull for possible use in later pipeline steps.

```{r}
#| echo: false
#| warning: false
#| message: false
#| cache: true

tickers <- c("SPY", "XLB", "XLC", "XLE", "XLF", "XLI",
             "XLK", "XLP", "XLRE", "XLV", "XLU", "XLY")

df <- tidyquant::tq_get(
  tickers,
  get  = "stock.prices",
  from = "2000-06-01",
  to   = "2026-03-31"
)
```

## Background Setup

```{r}
#| include: false

df.train <- df %>% dplyr::filter(date <= as.Date("2024-12-31"))
df.test  <- df %>% dplyr::filter(date >= as.Date("2025-01-01"))

etf.ret.train <- df.train %>%
  dplyr::group_by(symbol) %>%
  dplyr::mutate(ret = log(adjusted / dplyr::lag(adjusted))) %>%
  tidyr::drop_na(ret) %>%
  dplyr::ungroup()

etf.ret.test <- df.test %>%
  dplyr::group_by(symbol) %>%
  dplyr::mutate(ret = log(adjusted / dplyr::lag(adjusted))) %>%
  tidyr::drop_na(ret) %>%
  dplyr::ungroup()
```

## Step 1 - Clustering

Before we can test any pair for mean-reversion, we need to narrow 10 sector ETFs down to pairs that have a structural economic reason to move together.

Clustering on daily log returns does that. We use log returns rather than prices because prices drift upward over 24 years, which makes the raw distance between two price series meaningless as a similarity measure. Log returns strip away that drift and leave only co-movement, which is the actual signal we care about. **We also cluster on returns alone rather than mixing in macro variables like interest rates, oil prices, or GDP growth. Those factors are already embedded in how ETFs move every day.**

If XLE is sensitive to oil, that sensitivity shows up in its realized returns relative to other sectors. Adding macro variables explicitly would double-count information that is already in the data, while introducing noise from slow-moving, non-stationary series that distort the distance calculation. XLC and XLRE are excluded because both have under 10 years of training history, which is not enough to estimate stable long-run relationships.

```{r}
#| echo: false
#| warning: false
#| message: false

# --- Data prep ---
tickers.cluster <- c("SPY", "XLB", "XLE", "XLF", "XLI",
                     "XLK", "XLP", "XLV", "XLU", "XLY")

etf.wide.train <- etf.ret.train %>%
  dplyr::filter(symbol %in% tickers.cluster) %>%
  tidyr::pivot_wider(
    id_cols     = date,
    names_from  = symbol,
    values_from = ret
  ) %>%
  tidyr::drop_na()

recipe.cluster <- recipes::recipe(~ ., data = etf.wide.train) %>%
  recipes::step_rm(date) %>%
  recipes::step_normalize(recipes::all_numeric()) %>%
  recipes::prep(training = etf.wide.train)

etf.baked  <- recipes::bake(recipe.cluster, new_data = etf.wide.train)
etf.scaled <- base::as.matrix(etf.baked)
etf.t      <- base::t(etf.scaled)
dist.mat   <- stats::dist(etf.t)

# --- WSS for elbow ---
wss <- purrr::map_dbl(2:8, function(k) {
  km <- stats::kmeans(etf.t, centers = k, nstart = 25)
  km$tot.withinss
})
df.elbow <- dplyr::tibble(k = 2:8, wss = wss)

# --- K-Means (k = 3) ---
set.seed(452)
km.result <- stats::kmeans(etf.t, centers = 3, nstart = 25)

def.cluster <- km.result$cluster[["XLU"]]
cyc.cluster <- km.result$cluster[["XLE"]]
grw.cluster <- km.result$cluster[["XLK"]]

df.clusters <- dplyr::tibble(
  symbol  = base::names(km.result$cluster),
  cluster = km.result$cluster
) %>%
  dplyr::mutate(
    cluster.label = dplyr::case_when(
      cluster == def.cluster ~ "Defensive",
      cluster == cyc.cluster ~ "Value / Cyclical",
      cluster == grw.cluster ~ "Growth",
      TRUE ~ paste0("Cluster ", cluster)
    ),
    sector = dplyr::case_when(
      symbol == "SPY"  ~ "S&P 500 Benchmark",
      symbol == "XLB"  ~ "Materials",
      symbol == "XLE"  ~ "Energy",
      symbol == "XLF"  ~ "Financials",
      symbol == "XLI"  ~ "Industrials",
      symbol == "XLK"  ~ "Technology",
      symbol == "XLP"  ~ "Consumer Staples",
      symbol == "XLV"  ~ "Health Care",
      symbol == "XLU"  ~ "Utilities",
      symbol == "XLY"  ~ "Consumer Discretionary",
      TRUE ~ symbol
    )
  ) %>%
  dplyr::arrange(cluster.label, symbol)

# --- Hierarchical (Ward's) ---
hc.result <- stats::hclust(dist.mat, method = "ward.D2")

# --- Silhouette ---
sil     <- cluster::silhouette(km.result$cluster, dist.mat)
sil.avg <- base::mean(sil[, 3])

# --- Stability: pre vs post 2020 ---
etf.wide.pre2020 <- etf.ret.train %>%
  dplyr::filter(symbol %in% tickers.cluster, date < as.Date("2020-01-01")) %>%
  tidyr::pivot_wider(id_cols = date, names_from = symbol, values_from = ret) %>%
  tidyr::drop_na()

recipe.pre2020 <- recipes::recipe(~ ., data = etf.wide.pre2020) %>%
  recipes::step_rm(date) %>%
  recipes::step_normalize(recipes::all_numeric()) %>%
  recipes::prep(training = etf.wide.pre2020)

etf.pre2020 <- base::t(base::as.matrix(
  recipes::bake(recipe.pre2020, new_data = etf.wide.pre2020)
))

set.seed(452)
km.pre2020 <- stats::kmeans(etf.pre2020, centers = 3, nstart = 25)

etf.wide.post2020 <- etf.ret.train %>%
  dplyr::filter(symbol %in% tickers.cluster, date >= as.Date("2020-01-01")) %>%
  tidyr::pivot_wider(id_cols = date, names_from = symbol, values_from = ret) %>%
  tidyr::drop_na()

recipe.post2020 <- recipes::recipe(~ ., data = etf.wide.post2020) %>%
  recipes::step_rm(date) %>%
  recipes::step_normalize(recipes::all_numeric()) %>%
  recipes::prep(training = etf.wide.post2020)

etf.post2020 <- base::t(base::as.matrix(
  recipes::bake(recipe.post2020, new_data = etf.wide.post2020)
))

set.seed(452)
km.post2020 <- stats::kmeans(etf.post2020, centers = 3, nstart = 25)

def.pre  <- km.pre2020$cluster[["XLU"]]
cyc.pre  <- km.pre2020$cluster[["XLE"]]
grw.pre  <- km.pre2020$cluster[["XLK"]]
def.post <- km.post2020$cluster[["XLU"]]
cyc.post <- km.post2020$cluster[["XLE"]]
grw.post <- km.post2020$cluster[["XLK"]]

df.stability <- dplyr::tibble(
  symbol    = base::names(km.pre2020$cluster),
  pre.2020  = dplyr::case_when(
    km.pre2020$cluster == def.pre  ~ "Defensive",
    km.pre2020$cluster == cyc.pre  ~ "Value / Cyclical",
    km.pre2020$cluster == grw.pre  ~ "Growth",
    TRUE ~ "Other"
  ),
  post.2020 = dplyr::case_when(
    km.post2020$cluster == def.post  ~ "Defensive",
    km.post2020$cluster == cyc.post  ~ "Value / Cyclical",
    km.post2020$cluster == grw.post  ~ "Growth",
    TRUE ~ "Other"
  )
) %>%
  dplyr::mutate(stable = dplyr::if_else(pre.2020 == post.2020, "Yes", "No"))
```

### Cluster Assignments

Three groups emerge that map onto how institutional investors actually think about sector rotation.

-   Defensive sectors (XLP, XLU) hold their value in downturns and carry low market beta.

-   Value/Cyclical sectors (XLE, XLF) are driven by commodity prices and interest rate regimes, not the broad market.

-   The Growth cluster captures the remaining sectors that are technology-heavy and tightly tied to SPY.

**Both the elbow plot and the dendrogram support k = 3.** The elbow has no sharp kink, but that is expected: every sector ETF carries heavy SPY exposure, which compresses the distance between clusters.

```{r}
#| echo: false
#| warning: false
#| message: false

p.elbow <- df.elbow %>%
  ggplot2::ggplot(ggplot2::aes(x = k, y = wss)) +
  ggplot2::geom_line(col = "blue") +
  ggplot2::geom_point(col = "blue", size = 3) +
  ggplot2::geom_vline(xintercept = 3, col = "red", linetype = "dashed") +
  ggplot2::scale_x_continuous(breaks = 2:8) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title    = "Elbow Method — K-Means WSS by Number of Clusters",
    subtitle = "Red dashed line at k = 3 | Training data only: Jun 2000 – Dec 2024",
    x        = "Number of Clusters (k)",
    y        = "Total Within-Cluster SS"
  )

p.dend <- factoextra::fviz_dend(
  hc.result,
  k         = 3,
  rect      = TRUE,
  rect_fill = TRUE,
  ggtheme   = ggplot2::theme_minimal(),
  main      = "Dendrogram — Ward's Method (k = 3)",
  sub       = "Training period: Jun 2000 – Dec 2024",
  xlab      = "",
  ylab      = "Height"
)

p.elbow
p.dend

df.clusters %>%
  dplyr::select(symbol, sector, cluster.label) %>%
  dplyr::rename(
    Ticker  = symbol,
    Sector  = sector,
    Cluster = cluster.label
  ) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "K-Means Cluster Assignments (k = 3)",
    subtitle = "Fit on training data Jun 2000 – Dec 2024 | Seed = 452"
  ) %>%
  gt::tab_row_group(label = "Growth",          rows = Cluster == "Growth") %>%
  gt::tab_row_group(label = "Value / Cyclical", rows = Cluster == "Value / Cyclical") %>%
  gt::tab_row_group(label = "Defensive",        rows = Cluster == "Defensive") %>%
  gt::tab_footnote(
    footnote  = "XLC (Communication Services) and XLRE (Real Estate) excluded due to insufficient training history.",
    locations = gt::cells_title(groups = "subtitle")
  )
```

### Silhouette Validation

The average silhouette score is approximately 0.16, which is low but not a red flag. It reflects the structure of the data rather than a problem with the clustering. **Because every sector ETF carries beta close to 1.0 against SPY, the algorithm is looking for sub-structure inside a universe where everything is already highly correlated. Tight, well-separated clusters are not achievable here.**

The purpose of clustering is to nominate economically plausible pairs, not to find perfect separation. The statistical validation comes in Step 2 through cointegration testing.

```{r}
#| echo: false
#| warning: false
#| message: false

p.sil <- factoextra::fviz_silhouette(
  sil,
  ggtheme       = ggplot2::theme_minimal(),
  print.summary = FALSE
) +
  ggplot2::labs(
    title    = paste0(
      "Silhouette Plot — K-Means (k = 3) | Avg Score = ",
      base::round(sil.avg, 3)
    ),
    subtitle = "Low score expected: sector ETFs share systematic market exposure by construction"
  )

p.sil

dplyr::tibble(
  Metric = c("Average Silhouette Score", "Cluster Quality"),
  Value  = c(
    base::as.character(base::round(sil.avg, 3)),
    dplyr::case_when(
      sil.avg >= 0.50 ~ "Strong (>= 0.5) — proceed",
      sil.avg >= 0.30 ~ "Moderate (0.3 – 0.5) — acceptable",
      TRUE            ~ "Weak (< 0.3) — revisit k or features"
    )
  )
) %>%
  gt::gt() %>%
  gt::tab_header(title = "K-Means Cluster Validation — Silhouette") %>%
  gt::tab_style(
    style     = gt::cell_text(color = "red", weight = "bold"),
    locations = gt::cells_body(
      columns = Value,
      rows    = Metric == "Cluster Quality" & grepl("Weak", Value)
    )
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(color = "blue", weight = "bold"),
    locations = gt::cells_body(
      columns = Value,
      rows    = Metric == "Cluster Quality" & grepl("Strong|Moderate", Value)
    )
  )
```

------------------------------------------------------------------------

## Step 2 - Cointegration Testing

Clustering tells us which sectors have historically moved together, but economic similarity is not the same as a tradeable relationship. Two ETFs can be highly correlated over 24 years and still have a spread that drifts permanently in one direction. A persistent drift means every trade betting on reversion compounds losses over time. **We need statistical proof that the spread is bounded and mean-reverting before committing capital to any pair.**

That proof comes from cointegration testing. We follow the Engle-Granger two-step procedure as implemented by Yang, Huang and Chen (2023).

-   The first step runs an OLS regression of log(P_A) on log(P_B) to estimate the hedge ratio beta. This beta determines how many units of ETF B to hold short against one long unit of ETF A so the resulting spread has the smallest possible long-run drift.

-   The second step runs an Augmented Dickey-Fuller test on those OLS residuals, which are the spread itself. The ADF null hypothesis is that the spread has a unit root, meaning it can wander arbitrarily far from its mean and never return. Rejecting that null at p \< 0.05 confirms the spread is stationary and mean-reverting. The pair is cointegrated and proceeds to Step 3.

Any pair that fails is dropped entirely, because trading a non-stationary spread has no theoretical bound on the loss. Every within-cluster pair from Step 1 is tested.

```{r}
#| echo: false
#| warning: false
#| message: false

# Log-price wide matrices — CLOSE (used for cointegration, OU, Kalman, and CC return)
log.prices.wide.train <- df.train %>%
  dplyr::filter(symbol %in% tickers.cluster) %>%
  dplyr::select(date, symbol, adjusted) %>%
  tidyr::pivot_wider(names_from = symbol, values_from = adjusted) %>%
  dplyr::mutate(dplyr::across(where(is.numeric), log)) %>%
  tidyr::drop_na()

log.prices.wide.test <- df.test %>%
  dplyr::filter(symbol %in% tickers.cluster) %>%
  dplyr::select(date, symbol, adjusted) %>%
  tidyr::pivot_wider(names_from = symbol, values_from = adjusted) %>%
  dplyr::mutate(dplyr::across(where(is.numeric), log)) %>%
  tidyr::drop_na()

# Log-price wide matrices — OPEN (adjusted for splits/dividends via adj ratio)
# Adj ratio = adjusted / close; apply to open: open.adj = open * (adjusted / close)
log.open.wide.train <- df.train %>%
  dplyr::filter(symbol %in% tickers.cluster) %>%
  dplyr::mutate(open.adj = open * (adjusted / close)) %>%
  dplyr::select(date, symbol, open.adj) %>%
  tidyr::pivot_wider(names_from = symbol, values_from = open.adj) %>%
  dplyr::mutate(dplyr::across(where(is.numeric), log)) %>%
  tidyr::drop_na()

log.open.wide.test <- df.test %>%
  dplyr::filter(symbol %in% tickers.cluster) %>%
  dplyr::mutate(open.adj = open * (adjusted / close)) %>%
  dplyr::select(date, symbol, open.adj) %>%
  tidyr::pivot_wider(names_from = symbol, values_from = open.adj) %>%
  dplyr::mutate(dplyr::across(where(is.numeric), log)) %>%
  tidyr::drop_na()
```

### Candidate Pair Generation

All within-cluster pairs are generated from the K-Means assignments. Pairs from different clusters are excluded as they have no structural economic relationship.

```{r}
#| echo: false
#| warning: false
#| message: false

# Generate all within-cluster pairs (no duplicates, no self-pairs)
df.pairs <- df.clusters %>%
  dplyr::rename(symbolA = symbol, clusterA = cluster.label) %>%
  dplyr::inner_join(
    df.clusters %>% dplyr::rename(symbolB = symbol, clusterB = cluster.label),
    by           = c("clusterA" = "clusterB"),
    relationship = "many-to-many"
  ) %>%
  dplyr::filter(symbolA < symbolB) %>%
  dplyr::select(symbolA, symbolB, cluster = clusterA)
```

### ADF Test - Engle-Granger Cointegration

For each candidate pair (A, B), OLS is run on log prices: `log(P_A) ~ log(P_B)`. The ADF test on the residuals determines whether the spread is stationary. Pairs with p-value \< 0.05 are cointegrated and proceed.

```{r}
#| echo: false
#| warning: false
#| message: false

df.coint <- df.pairs %>%
  dplyr::mutate(
    beta.ols = purrr::map2_dbl(symbolA, symbolB, function(a, b) {
      fit <- stats::lm(
        log.prices.wide.train[[a]] ~ log.prices.wide.train[[b]]
      )
      broom::tidy(fit)$estimate[2]
    }),
    adf.stat = purrr::map2_dbl(symbolA, symbolB, function(a, b) {
      fit   <- stats::lm(
        log.prices.wide.train[[a]] ~ log.prices.wide.train[[b]]
      )
      resid <- stats::resid(fit)
      base::suppressWarnings(tseries::adf.test(resid))$statistic
    }),
    adf.pval = purrr::map2_dbl(symbolA, symbolB, function(a, b) {
      fit   <- stats::lm(
        log.prices.wide.train[[a]] ~ log.prices.wide.train[[b]]
      )
      resid <- stats::resid(fit)
      base::suppressWarnings(tseries::adf.test(resid))$p.value
    }),
    cointegrated = dplyr::if_else(adf.pval < 0.05, "Yes", "No")
  )
```

```{r}
#| echo: false
#| warning: false
#| message: false

df.coint %>%
  dplyr::arrange(adf.pval) %>%
  dplyr::select(symbolA, symbolB, cluster, beta.ols, adf.stat, adf.pval, cointegrated) %>%
  dplyr::rename(
    `ETF A`        = symbolA,
    `ETF B`        = symbolB,
    Cluster        = cluster,
    `OLS Beta`     = beta.ols,
    `ADF Stat`     = adf.stat,
    `ADF p-value`  = adf.pval,
    Cointegrated   = cointegrated
  ) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Cointegration Test Results — All Candidate Pairs",
    subtitle = "Engle-Granger ADF test on OLS residuals | Training period: Jun 2000 – Dec 2024"
  ) %>%
  gt::fmt_number(columns = c(`OLS Beta`, `ADF Stat`, `ADF p-value`), decimals = 4) %>%
  gt::tab_style(
    style     = gt::cell_text(color = "blue", weight = "bold"),
    locations = gt::cells_body(
      columns = Cointegrated,
      rows    = Cointegrated == "Yes"
    )
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(color = "red"),
    locations = gt::cells_body(
      columns = Cointegrated,
      rows    = Cointegrated == "No"
    )
  ) %>%
  gt::tab_footnote(
    footnote  = "Null hypothesis: series has unit root (non-stationary). Reject at p < 0.05 → spread is stationary → pair is cointegrated.",
    locations = gt::cells_column_labels(columns = `ADF p-value`)
  )
```

### Cointegration Results

The test narrows the 22 within-cluster candidate pairs down to a smaller set of cointegrated survivors.

Most of the failures come from the Defensive and Value/Cyclical clusters, where idiosyncratic drivers such as recession demand and commodity price cycles make it difficult for spreads to stay bounded over 24 years. **The pairs that survive come from the Growth cluster, where shared technology exposure and market-cap concentration create tighter structural linkages that persist across full business cycles.**

------------------------------------------------------------------------

## Step 3 - OU Half-Life

Cointegration confirms that a mean-reverting spread exists, but it says nothing about how quickly it reverts.

A pair with a half-life of 240 days is statistically cointegrated and practically useless in the same breath. Holding a position for nearly a full year ties up capital, accumulates financing costs, and gives the underlying relationship more time to structurally break down mid-trade. We need to rank the surviving pairs by reversion speed and select the one that converts a statistical finding into an actionable holding period.

We used the Ornstein-Uhlenbeck process to models the spread as a continuous mean-reverting system. The reversion speed is estimated by regressing the daily change in the spread on the prior day's spread level. The coefficient on the lagged spread, called theta, captures the pull-back force toward the mean. A more negative theta means stronger and faster reversion. The half-life is then -ln(2) / theta, which converts that coefficient into a number of days.

Any pair where theta is non-negative has no mean-reverting pull and is discarded. From the remaining **pairs, a ceiling of 252 trading days is applied.** This is more generous than the 60-day cutoff commonly used in individual equity pairs trading, because sector ETFs are diversified portfolios with idiosyncratic risk already averaged out. Their residual mean-reversion operates on longer timescales by construction.

```{r}
#| echo: false
#| warning: false
#| message: false

df.valid <- df.coint %>%
  dplyr::filter(cointegrated == "Yes") %>%
  dplyr::mutate(
    half.life = purrr::pmap_dbl(
      list(symbolA, symbolB, beta.ols),
      function(a, b, hr) {
        spread <- log.prices.wide.train[[a]] - hr * log.prices.wide.train[[b]]
        df.ou  <- dplyr::tibble(
          spread     = spread,
          spread.lag = dplyr::lag(spread),
          delta      = spread - spread.lag
        ) %>% tidyr::drop_na()
        fit.ou <- stats::lm(delta ~ spread.lag, data = df.ou)
        theta  <- broom::tidy(fit.ou)$estimate[2]
        -base::log(2) / theta
      }
    ),
    tradeable = dplyr::if_else(
      half.life > 0 & half.life <= 252, "Yes", "No"
    )
  ) %>%
  dplyr::arrange(half.life)
```

```{r}
#| echo: false
#| warning: false
#| message: false

df.valid %>%
  dplyr::select(symbolA, symbolB, cluster, beta.ols, adf.pval, half.life, tradeable) %>%
  dplyr::rename(
    `ETF A`       = symbolA,
    `ETF B`       = symbolB,
    Cluster       = cluster,
    `OLS Beta`    = beta.ols,
    `ADF p-value` = adf.pval,
    `Half-Life (days)` = half.life,
    Tradeable     = tradeable
  ) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "OU Half-Life — Cointegrated Pairs",
    subtitle = "Short half-life = faster mean reversion = more tradeable | Discard HL > 252 days or θ ≥ 0"
  ) %>%
  gt::fmt_number(
    columns  = c(`OLS Beta`, `ADF p-value`, `Half-Life (days)`),
    decimals = 3
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(color = "blue", weight = "bold"),
    locations = gt::cells_body(
      columns = Tradeable,
      rows    = Tradeable == "Yes"
    )
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(color = "red"),
    locations = gt::cells_body(
      columns = Tradeable,
      rows    = Tradeable == "No"
    )
  )
```

```{r}
#| include: false

# All tradeable pairs — traded simultaneously with equal weight
pairs.tradeable <- df.valid %>%
  dplyr::filter(tradeable == "Yes") %>%
  dplyr::arrange(half.life)

# Primary pair (shortest half-life) — retained as reference label for CC/CO/OC diagnostics
tickerA <- pairs.tradeable$symbolA[1]
tickerB <- pairs.tradeable$symbolB[1]
```

All `r nrow(pairs.tradeable)` tradeable pairs proceed to the strategy layer and are traded simultaneously with equal weight allocation, rebalanced to equal weight at the end of each calendar month. The fastest-reverting pair is **`r tickerA` / `r tickerB`**, which serves as the reference pair for execution timing diagnostics in Step 4.

### Half-Life Results

All tradeable pairs come from the Growth cluster. The Defensive and Value/Cyclical pairs were eliminated in Step 2, which confirms that structural economic linkage does not guarantee a tradeable spread. Growth ETFs dominate because shared technology exposure and large-cap concentration keep their spreads from drifting permanently across business cycles.

The fastest-reverting pair is XLV/XLY (Health Care vs Consumer Discretionary). Both are large-cap domestic sectors with similar overall market sensitivity, but they respond to opposite ends of the consumer cycle. Health Care spending is non-discretionary and holds up in downturns. Consumer Discretionary contracts when confidence falls. Because neither sector permanently outpaces the other over a full business cycle, the spread oscillates around a stable mean. Rather than concentrating the entire strategy in one pair, all tradeable pairs are traded simultaneously. Each pair contributes an independent source of mean-reversion alpha, and monthly rebalancing to equal weight systematically sells pairs that have outperformed and buys those that have underperformed, compounding the mean-reversion effect at the portfolio level.

------------------------------------------------------------------------

## Step 4 - Kalman Filter and Strategy

### Strategy Function

Cointegration confirmed that the spread between two ETFs is stationary, it cannot drift permanently. But to actually trade it, we need to know how many units of ETF B to short against one long unit of ETF A at any given time. That ratio is the hedge ratio β.

The naive approach is to reuse the OLS beta from Step 2. The problem: that beta is a single point estimate fit on 24 years of price history. Sector relationships shift across business cycles. The same two ETFs that moved dollar-for-dollar in 2005 may have drifted to a 1.2:1 relationship by 2015 as their compositions and market roles evolved. A static hedge ratio will progressively mismatch the spread, generating signals against a spread that no longer means what it used to.

**The Kalman filter solves this.** Introduced by Kalman (1960) and applied to pairs trading by Avellaneda & Lee (2008) and Yang, Huang & Chen (2023), it is a recursive Bayesian estimator that updates β_t every single day.

Mechanically :

-   It treats the true hedge ratio as a hidden state that evolves over time. Each new price observation updates the estimate by blending the prior day's value with a correction term scaled by the Kalman gain K_t. The gain is governed by the ratio Q/H, process noise to observation noise. High Q/H means the hedge ratio is volatile, so trust the new data more. Low Q/H means the hedge ratio is stable, so stay close to the prior.

-   The calibration follows Yang, Huang & Chen (2023): Q = α² / (1 − α) × H, where α = 2 / (n.ma + 1). This ties the filter's adaptation speed directly to a target moving-average window n.ma. A larger n.ma produces a smoother, slower-adapting β_t — appropriate for sector pairs that mean-revert over months. The optimal n.ma is determined by the grid search in Section 8.7.

-   Once β_t is known, the spread is: spread_t = log(P_A,t) − β_t × log(P_B,t). The z-score of this spread against its 20-day rolling mean and standard deviation is the trade signal. Z \> +z.entry means the spread is stretched, short it. Z \< −z.entry means the spread is compressed, long it. The position closes when z returns to zero, or is stopped at ±z.stop.

The full strategy logic is encapsulated in two functions. `strategy_pair()` runs the Kalman filter and signal generation for a single pair. `portfolio_pairs()` wraps it across all tradeable pairs, aggregates their daily returns with equal weight allocation, and rebalances to equal weight at the end of each calendar month.

```{r}
#| echo: false
#| warning: false
#| message: false

strategy_pair <- function(prices.wide,
                          tickerA,
                          tickerB,
                          n.ma       = 60,
                          z.entry    = 2,
                          z.stop     = 3,
                          open.wide  = NULL) {

  alpha <- 2 / (n.ma + 1)
  q     <- alpha^2 / (1 - alpha)

  vec.A  <- prices.wide[[tickerA]]
  vec.B  <- prices.wide[[tickerB]]
  n.obs  <- base::length(vec.A)

  H.est  <- stats::var(vec.A - vec.B, na.rm = TRUE)
  Q.est  <- q * H.est

  # Kalman filter: state = hedge ratio beta_t
  beta_t <- base::numeric(n.obs)
  P.filt <- H.est

  for (i in base::seq_len(n.obs)) {
    P.pred    <- P.filt + Q.est
    Bi        <- vec.B[i]
    K         <- P.pred * Bi / (Bi^2 * P.pred + H.est)
    beta.pred <- if (i == 1L) 1 else beta_t[i - 1L]
    beta_t[i] <- beta.pred + K * (vec.A[i] - Bi * beta.pred)
    P.filt    <- (1 - K * Bi) * P.pred
  }

  result <- prices.wide %>%
    dplyr::mutate(
      beta_t  = beta_t,
      spread  = .data[[tickerA]] - beta_t * .data[[tickerB]],
      z.mean  = slider::slide_dbl(spread, mean,       .before = 19, .after = 0, .complete = FALSE),
      z.sd    = slider::slide_dbl(spread, stats::sd,  .before = 19, .after = 0, .complete = FALSE),
      z.score = (spread - z.mean) / z.sd,
      # Close-to-close log returns per leg
      ret.A   = .data[[tickerA]] - dplyr::lag(.data[[tickerA]]),
      ret.B   = .data[[tickerB]] - dplyr::lag(.data[[tickerB]]),
      signal  = dplyr::case_when(
        base::abs(z.score) > z.stop ~ 0L,
        z.score >  z.entry           ~ -1L,
        z.score < -z.entry           ~ 1L,
        TRUE                         ~ 0L
      ),
      trade   = tidyr::replace_na(
        dplyr::lag(signal) - dplyr::lag(signal, n = 2L), 0L
      ),
      pos          = base::cumsum(trade),
      prev.pos     = dplyr::lag(pos, default = 0L),
      # Close → Close spread return
      spread.ret.cc = ret.A - beta_t * ret.B
    )

  if (!base::is.null(open.wide)) {
    # Compute intraday (OC) and overnight (CO) spread returns
    result <- result %>%
      dplyr::mutate(
        open.A        = open.wide[[tickerA]],
        open.B        = open.wide[[tickerB]],
        ret.co.A      = open.A - dplyr::lag(.data[[tickerA]]),
        ret.co.B      = open.B - dplyr::lag(.data[[tickerB]]),
        spread.ret.co = ret.co.A - beta_t * ret.co.B,
        ret.oc.A      = .data[[tickerA]] - open.A,
        ret.oc.B      = .data[[tickerB]] - open.B,
        spread.ret.oc = ret.oc.A - beta_t * ret.oc.B
      ) %>%
      dplyr::mutate(
        # Entry day (prev flat → in trade): fill at open → earn OC
        # Exit day  (in trade → flat):      exit at open → earn CO on prior leg
        # Hold / flip:                       earn full CC
        ret = dplyr::case_when(
          prev.pos == 0L & pos != 0L ~ pos      * tidyr::replace_na(spread.ret.oc, 0),
          pos == 0L & prev.pos != 0L ~ prev.pos * tidyr::replace_na(spread.ret.co, 0),
          TRUE                       ~ pos      * tidyr::replace_na(spread.ret.cc, 0)
        ),
        cumeq    = base::cumprod(1 + ret),
        ret.co   = pos * tidyr::replace_na(spread.ret.co, 0),
        cumeq.co = base::cumprod(1 + ret.co),
        ret.oc   = pos * tidyr::replace_na(spread.ret.oc, 0),
        cumeq.oc = base::cumprod(1 + ret.oc)
      )
  } else {
    result <- result %>%
      dplyr::mutate(
        ret   = pos * tidyr::replace_na(spread.ret.cc, 0),
        cumeq = base::cumprod(1 + ret)
      )
  }

  result
}
```

```{r}
#| include: false

# portfolio_pairs(): runs strategy_pair() for every row of pairs.df,
# aggregates to an equal-weight portfolio with monthly rebalancing
portfolio_pairs <- function(pairs.df,
                            prices.wide,
                            open.wide  = NULL,
                            n.ma       = 60,
                            z.entry    = 2,
                            z.stop     = 3) {

  # Run strategy_pair for each pair; keep only date + daily CC return
  pair.results <- purrr::pmap(
    list(pairs.df$symbolA, pairs.df$symbolB),
    function(a, b) {
      strategy_pair(
        prices.wide = prices.wide,
        tickerA     = a,
        tickerB     = b,
        n.ma        = n.ma,
        z.entry     = z.entry,
        z.stop      = z.stop,
        open.wide   = open.wide   # pass through for realistic execution timing
      ) %>%
        dplyr::select(date, ret) %>%
        dplyr::rename(!!paste0(a, "_", b) := ret)
    }
  )

  # Align all pair return streams on the same dates
  df.wide   <- purrr::reduce(pair.results, dplyr::inner_join, by = "date") %>%
    tidyr::drop_na()
  pair.cols <- base::setdiff(base::names(df.wide), "date")
  n.p       <- base::length(pair.cols)

  # Equal-weight allocation tracker (normalised to start at 1)
  allocs             <- base::rep(1 / n.p, n.p)
  base::names(allocs) <- pair.cols

  n.rows   <- base::nrow(df.wide)
  port.val <- base::numeric(n.rows)
  port.ret <- base::numeric(n.rows)
  port.val[1L] <- 1

  for (i in 2L:n.rows) {
    # Grow each allocation by its return
    for (col in pair.cols) {
      allocs[col] <- allocs[col] * (1 + df.wide[[col]][i])
    }
    port.val[i] <- base::sum(allocs)
    port.ret[i] <- port.val[i] / port.val[i - 1L] - 1

    # Monthly rebalance: detect month boundary between day i and day i+1
    if (i < n.rows) {
      curr.m <- lubridate::floor_date(df.wide$date[i],        "month")
      next.m <- lubridate::floor_date(df.wide$date[i + 1L],   "month")
      if (next.m > curr.m) {
        allocs             <- base::rep(port.val[i] / n.p, n.p)
        base::names(allocs) <- pair.cols
      }
    }
  }

  df.wide %>%
    dplyr::mutate(
      ret   = port.ret,
      cumeq = base::cumprod(1 + ret)
    ) %>%
    dplyr::select(date, dplyr::all_of(pair.cols), ret, cumeq)
}
```

```{r}
#| include: false

# Reference pair: run with open prices for CC/CO/OC execution timing diagnostics
result.train <- strategy_pair(
  prices.wide = log.prices.wide.train,
  tickerA     = tickerA,
  tickerB     = tickerB,
  n.ma        = 60,
  z.entry     = 2,
  z.stop      = 3,
  open.wide   = log.open.wide.train
)
```

```{r}
#| include: false

# Full portfolio: all tradeable pairs, equal weight, monthly rebalancing
port.train <- portfolio_pairs(
  pairs.df    = pairs.tradeable,
  prices.wide = log.prices.wide.train,
  open.wide   = log.open.wide.train,
  n.ma        = 60,
  z.entry     = 2,
  z.stop      = 3
)
```

### Execution Timing

The strategy submits market-on-open (MOO) orders the day after a z-score signal fires. This avoids any look-ahead from the closing price that generated the signal. The return is computed using the appropriate convention per day:

-   **Entry day** - position opens at the next-day open; spread earns from open to close (OC)
-   **Holding days** - fully in position; spread earns the full close-to-close return (CC)
-   **Exit day** - position closes at the next-day open; spread earns only the overnight gap (CO)

```{r}
#| include: false

df.exec <- result.train %>%
  tidyr::drop_na(cumeq, cumeq.co, cumeq.oc) %>%
  dplyr::select(date, CC = cumeq, CO = cumeq.co, OC = cumeq.oc) %>%
  tidyr::pivot_longer(cols = c(CC, CO, OC), names_to = "Convention", values_to = "CumEQ")
```

```{r}
#| include: false

dplyr::tibble(
  Convention = c("Close → Close (CC)", "Close → Open (CO)", "Open → Close (OC)"),
  `Cum. Return` = c(
    base::round(dplyr::last(result.train$cumeq) - 1, 4),
    base::round(dplyr::last(result.train$cumeq.co) - 1, 4),
    base::round(dplyr::last(result.train$cumeq.oc) - 1, 4)
  )
) %>% gt::gt()
```

### Portfolio Equity - Default Parameters

```{r}
#| include: false

# Single-pair panels (hidden — kept for internal reference)
p1 <- result.train %>%
  tidyr::drop_na(z.score) %>%
  ggplot2::ggplot(ggplot2::aes(x = date, y = z.score)) +
  ggplot2::geom_line(col = "blue", linewidth = 0.3) +
  ggplot2::geom_hline(yintercept =  2, col = "red", linetype = "dashed") +
  ggplot2::geom_hline(yintercept = -2, col = "red", linetype = "dashed") +
  ggplot2::geom_hline(yintercept =  0, col = "grey20") +
  ggplot2::theme_minimal() +
  ggplot2::labs(title = paste0(tickerA, " / ", tickerB, " — Reference Pair"), x = "", y = "Z-Score")

p2 <- result.train %>%
  tidyr::drop_na(trade) %>%
  ggplot2::ggplot(ggplot2::aes(x = date, y = trade)) +
  ggplot2::geom_col(fill = "blue") +
  ggplot2::theme_minimal() +
  ggplot2::labs(x = "", y = "Trades")

p3 <- result.train %>%
  tidyr::drop_na(pos) %>%
  ggplot2::ggplot(ggplot2::aes(x = date, y = pos)) +
  ggplot2::geom_step(col = "blue") +
  ggplot2::geom_hline(yintercept = 0, col = "grey20") +
  ggplot2::theme_minimal() +
  ggplot2::labs(x = "", y = "Position")

p4 <- result.train %>%
  tidyr::drop_na(cumeq) %>%
  ggplot2::ggplot(ggplot2::aes(x = date, y = cumeq)) +
  ggplot2::geom_line(col = "blue") +
  ggplot2::geom_hline(yintercept = 1, col = "grey20") +
  ggplot2::theme_minimal() +
  ggplot2::labs(x = "", y = "CumEQ")
```

```{r}
#| echo: false
#| warning: false
#| message: false

# Portfolio equity curve — all tradeable pairs, equal weight, monthly rebalancing
port.train %>%
  tidyr::drop_na(cumeq) %>%
  ggplot2::ggplot(ggplot2::aes(x = date, y = cumeq)) +
  ggplot2::geom_line(col = "blue", linewidth = 0.8) +
  ggplot2::geom_hline(yintercept = 1, col = "grey20") +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title    = "Portfolio Equity — All Tradeable Pairs (Default Parameters)",
    subtitle = paste0(
      nrow(pairs.tradeable), " pairs | Equal weight | Monthly rebalancing",
      " | z.entry = 2, n.ma = 60 | Training: Jun 2000 – Dec 2024"
    ),
    x = "",
    y = "Cumulative Equity"
  )
```

**Interpretation - Portfolio Equity (Default Parameters):**

The portfolio combines all tradeable pairs into a single equity stream. Each pair contributes an independent mean-reversion signal, and monthly rebalancing continuously redirects capital toward pairs with room to revert and away from pairs that have already closed their spread. The default parameters (z.entry = 2, n.ma = 60) are conservative, the strategy is in the market only when the spread is dislocated by at least two standard deviations. The optimization section finds the parameter combination that maximizes the risk-adjusted return across the full portfolio.

------------------------------------------------------------------------

## Optimization

The strategy has two tunable parameters: **`z.entry`** (signal threshold) and **`n.ma`** (Kalman adaptation window). In this section, we will optimize the two parameters in order to get the best risk to reward ratio.

### Strategy Function and Parameter Grid

```{r}
#| echo: false
#| warning: false
#| message: false
#| cache: true

# Parameter grid (8.7.4.1: expand.grid approach) — fine mesh for 3D surface (8.8.9)
out <- base::expand.grid(
  z.entry = base::seq(0.50, 2.75, by = 0.25),
  n.ma    = base::seq(10L, 120L, by = 10L)
)

# Sweep: purrr::map2 — vectorize, not for-loops
# Each combination is evaluated on the full equal-weight portfolio
safe_trade_stats <- purrr::possibly(
  function(ze, nm) {
    portfolio_pairs(
      pairs.df    = pairs.tradeable,
      prices.wide = log.prices.wide.train,
      open.wide   = log.open.wide.train,
      n.ma        = nm,
      z.entry     = ze,
      z.stop      = 3
    ) %>%
      dplyr::select(date, ret) %>%
      RTL::tradeStats()
  },
  otherwise = NULL
)

res <- out %>%
  dplyr::mutate(
    stats = purrr::map2(z.entry, n.ma, safe_trade_stats)
  ) %>%
  dplyr::filter(!purrr::map_lgl(stats, is.null)) %>%
  tidyr::unnest_wider(stats)
```

### Optimization Visualization

```{r}
#| echo: false
#| warning: false
#| message: false

make_heatmap <- function(df, metric_col, title_str) {
  df %>%
    dplyr::select(z.entry, n.ma, value = dplyr::all_of(metric_col)) %>%
    dplyr::mutate(
      z.entry = base::factor(z.entry),
      n.ma    = base::factor(n.ma)
    ) %>%
    ggplot2::ggplot(ggplot2::aes(x = z.entry, y = n.ma, fill = value)) +
    ggplot2::geom_tile(col = "white") +
    ggplot2::scale_fill_gradient2(
      low      = "red",
      mid      = "white",
      high     = "blue",
      midpoint = 0
    ) +
    ggplot2::theme_minimal() +
    ggplot2::labs(
      title = title_str,
      x     = "Z-Score Entry Threshold",
      y     = "Kalman MA Window (n.ma)",
      fill  = ""
    )
}

p.sharpe  <- make_heatmap(res, "Sharpe",    "Sharpe Ratio")
p.cumret  <- make_heatmap(res, "CumReturn", "Cumulative Return")
p.ddmax   <- make_heatmap(res, "DD.Max",    "Max Drawdown")

(p.sharpe | p.cumret) / p.ddmax
```

**Interpretation - Optimization Heatmaps:**

The heatmaps show how the strategy's performance varies across the full parameter grid. Reading them together:

-   **Sharpe Ratio**: The highest Sharpe values cluster in the low `z.entry` and moderate-to-high `n.ma` region. Low z.entry means the strategy enters trades more frequently (smaller dislocation required). Higher n.ma means a smoother Kalman filter that adapts slowly — appropriate for a slow-reverting pair.
-   **Cumulative Return**: Mirrors the Sharpe pattern, more frequent trading (low z.entry) generates higher total return.
-   **Max Drawdown**: More aggressive parameters (lower z.entry) also produce deeper drawdowns. This is the fundamental risk-return tradeoff.

**Key insight:** The heatmaps do **not** show a single dominant corner, there is no "free lunch" parameter. Higher return always comes with higher drawdown. The 3D surfaces below make the tradeoff structure explicit.

### 3D Surface Visualization

Three interactive surfaces show the optimization landscape. We are looking for **a broad plateau** (flat-but-high region), not a sharp spike. A spike means the parameter is fragile; a plateau means nearby parameters produce similar performance, which is the hallmark of robustness.

```{r}
#| echo: false
#| warning: false
#| message: false

par1.vals <- base::sort(base::unique(res$z.entry))
par2.vals <- base::sort(base::unique(res$n.ma))

mat.sharpe <- res %>%
  dplyr::select(z.entry, n.ma, Sharpe) %>%
  tidyr::pivot_wider(names_from = n.ma, values_from = Sharpe) %>%
  dplyr::select(-z.entry) %>%
  base::as.matrix()

mat.cumret <- res %>%
  dplyr::select(z.entry, n.ma, CumReturn) %>%
  tidyr::pivot_wider(names_from = n.ma, values_from = CumReturn) %>%
  dplyr::select(-z.entry) %>%
  base::as.matrix()

mat.ddmax <- res %>%
  dplyr::select(z.entry, n.ma, DD.Max) %>%
  tidyr::pivot_wider(names_from = n.ma, values_from = DD.Max) %>%
  dplyr::select(-z.entry) %>%
  base::as.matrix()

plotly::plot_ly(x = ~par2.vals, y = ~par1.vals, z = ~mat.sharpe) %>%
  plotly::add_surface(colorscale = "Viridis") %>%
  plotly::layout(
    title = list(text = "Sharpe Ratio Surface"),
    scene = list(
      xaxis = list(title = "n.ma (Kalman Window)"),
      yaxis = list(title = "z.entry (Threshold)"),
      zaxis = list(title = "Sharpe")
    )
  )
```

```{r}
#| echo: false
#| warning: false
#| message: false

plotly::plot_ly(x = ~par2.vals, y = ~par1.vals, z = ~mat.cumret) %>%
  plotly::add_surface(colorscale = "Viridis") %>%
  plotly::layout(
    title = list(text = "Cumulative Return Surface"),
    scene = list(
      xaxis = list(title = "n.ma (Kalman Window)"),
      yaxis = list(title = "z.entry (Threshold)"),
      zaxis = list(title = "CumReturn")
    )
  )
```

```{r}
#| echo: false
#| warning: false
#| message: false

plotly::plot_ly(x = ~par2.vals, y = ~par1.vals, z = ~mat.ddmax) %>%
  plotly::add_surface(colorscale = "Viridis") %>%
  plotly::layout(
    title = list(text = "Maximum Drawdown Surface"),
    scene = list(
      xaxis = list(title = "n.ma (Kalman Window)"),
      yaxis = list(title = "z.entry (Threshold)"),
      zaxis = list(title = "DD.Max")
    )
  )
```

**Reading the 3D surfaces:**

-   **Sharpe surface**: Look for the broad, elevated region. If the peak is a sharp needle, those parameters are overfit. If the peak forms a wide mesa, those parameters are robust.
-   **Cumulative Return surface**: Shows total return across the grid. A rising slope along one axis means that axis dominates performance.
-   **Max Drawdown surface**: Higher values (closer to zero) are better. Deep valleys indicate parameter combinations where the strategy suffered severe losses.

The "sweet spot" is the parameter region that is simultaneously **high Sharpe, high CumReturn, and shallow drawdown**, which is a bit hard to formalize only looking at the above charts, that is why the next section will solve this using Z-score normalization.

### Z-Score Normalization

To compare metrics on the same scale, each performance dimension is standardized to a Z-score: (Z = \frac{x - \mu}{\sigma}). A positive Z-score means "better than average across the grid." The composite Z-score is the sum of all three: a parameter combination that is above-average on *every* dimension receives a high composite score.

```{r}
#| echo: false
#| warning: false
#| message: false

res.z <- res %>%
  dplyr::select(z.entry, n.ma, Sharpe, CumReturn, DD.Max) %>%
  tidyr::pivot_longer(
    cols      = c(Sharpe, CumReturn, DD.Max),
    names_to  = "variable",
    values_to = "value"
  ) %>%
  dplyr::group_by(variable) %>%
  dplyr::mutate(
    value.z = (value - base::mean(value, na.rm = TRUE)) /
              stats::sd(value, na.rm = TRUE)
  ) %>%
  dplyr::ungroup()

res.z %>%
  ggplot2::ggplot(ggplot2::aes(x = value.z)) +
  ggplot2::geom_histogram(
    col  = "black",
    fill = "blue",
    ggplot2::aes(y = ggplot2::after_stat(density)),
    bins = 25
  ) +
  ggplot2::facet_wrap(~ variable, scales = "free_y") +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title    = "8.8.9.1 Z-Score Distributions by Metric",
    subtitle = "Positive Z = above-average performance | Used for composite robustness scoring",
    x        = "Z-Score",
    y        = "Density"
  )
```

### Robust Parameter Selection - Plateau Method

**Method:** For each grid point, compute a composite Z-score (sum of Z-scored Sharpe + CumReturn + DD.Max). Then compute the average composite of its immediate neighbors (±1 grid step). The **robustness score** is the minimum of the two: `robustness = min(own composite, neighbor average)`. This max-min criterion penalizes isolated spikes and rewards broad plateaus.

```{r}
#| echo: false
#| warning: false
#| message: false

composite <- res.z %>%
  dplyr::group_by(z.entry, n.ma) %>%
  dplyr::summarise(composite.z = base::sum(value.z), .groups = "drop")

z.step  <- 0.25
ma.step <- 10L

composite <- composite %>%
  dplyr::mutate(
    neighbor.avg = purrr::map2_dbl(z.entry, n.ma, function(ze, nm) {
      neighbors <- composite %>%
        dplyr::filter(
          base::abs(z.entry - ze) <= z.step,
          base::abs(n.ma - nm)    <= ma.step,
          !(z.entry == ze & n.ma == nm)
        )
      if (base::nrow(neighbors) == 0L) return(NA_real_)
      base::mean(neighbors$composite.z, na.rm = TRUE)
    }),
    robustness = base::pmin(composite.z, neighbor.avg)
  )

robust.pick <- composite %>%
  dplyr::filter(!base::is.na(robustness)) %>%
  dplyr::arrange(dplyr::desc(robustness)) %>%
  dplyr::slice(1)
```

```{r}
#| echo: false
#| warning: false
#| message: false

# Composite surface — the plateau is visible as a broad elevated region
mat.composite <- composite %>%
  dplyr::select(z.entry, n.ma, composite.z) %>%
  tidyr::pivot_wider(names_from = n.ma, values_from = composite.z) %>%
  dplyr::select(-z.entry) %>%
  base::as.matrix()

plotly::plot_ly(x = ~par2.vals, y = ~par1.vals, z = ~mat.composite) %>%
  plotly::add_surface(colorscale = "Viridis") %>%
  plotly::layout(
    title = list(text = "Composite Z-Score Surface (Sharpe + CumReturn + DD.Max)"),
    scene = list(
      xaxis = list(title = "n.ma"),
      yaxis = list(title = "z.entry"),
      zaxis = list(title = "Composite Z")
    )
  )
```

```{r}
#| echo: false
#| warning: false
#| message: false

# Highlight the robust pick and top-5 plateau points
composite %>%
  dplyr::filter(!base::is.na(robustness)) %>%
  dplyr::arrange(dplyr::desc(robustness)) %>%
  dplyr::slice(1:5) %>%
  dplyr::left_join(
    res %>% dplyr::select(z.entry, n.ma, Sharpe, CumReturn, DD.Max),
    by = c("z.entry", "n.ma")
  ) %>%
  dplyr::select(z.entry, n.ma, Sharpe, CumReturn, DD.Max, composite.z, robustness) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "8.8.9.2 Top-5 Robust Parameter Combinations",
    subtitle = "Ranked by min(own composite, neighbor avg) — plateau criterion"
  ) %>%
  gt::fmt_number(
    columns  = c(Sharpe, CumReturn, DD.Max, composite.z, robustness),
    decimals = 3
  ) %>%
  gt::tab_style(
    style     = gt::cell_fill(color = "#E8F5E9"),
    locations = gt::cells_body(rows = 1)
  ) %>%
  gt::tab_footnote(
    footnote  = "Green row = selected parameters. Plateau method guards against overfitting: isolated spikes are penalized.",
    locations = gt::cells_title(groups = "subtitle")
  )
```

**Interpretation - Robust Selection:**

The composite Z-score surface combines Sharpe, cumulative return, and drawdown into a single "desirability" metric. The plateau method selects from the broad elevated region rather than the absolute peak. **This is critical for out-of-sample stability: if the surface is jagged and we pick the tallest spike, we are likely fitting noise. If we pick from a mesa, nearby parameters would have produced similar results a.k.a the strategy is not fragile to small perturbations in parameter choice.**

### Risk Appetite - Select Optimal Parameters

Following the plateau robustness analysis above: the selected parameters are the most robust combination from the composite Z-score surface. As a secondary check, they must also satisfy a risk appetite floor (Sharpe \> 0 and Max Drawdown \> −0.50).

```{r}
#| echo: false
#| warning: false
#| message: false

# Use robust plateau pick from 8.8.9.2
optimal <- res %>%
  dplyr::filter(
    z.entry == robust.pick$z.entry,
    n.ma    == robust.pick$n.ma
  ) %>%
  dplyr::slice(1)

optimal %>%
  dplyr::select(z.entry, n.ma, CumReturn, Ret.Ann, SD.Ann, Sharpe, Omega, `%.Win`, DD.Max) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "8.9 Selected Parameters — Risk Appetite Filter",
    subtitle = "Selected via plateau robustness method (8.8.9.2) — flat-but-high across all 3 metrics"
  ) %>%
  gt::fmt_number(
    columns  = c(CumReturn, Ret.Ann, SD.Ann, Sharpe, Omega, `%.Win`, DD.Max),
    decimals = 4
  )
```

```{r}
#| include: false

# Re-run full portfolio with selected optimal parameters
result.opt <- portfolio_pairs(
  pairs.df    = pairs.tradeable,
  prices.wide = log.prices.wide.train,
  open.wide   = log.open.wide.train,
  n.ma        = optimal$n.ma,
  z.entry     = optimal$z.entry,
  z.stop      = 3
)
```

## Risk / Reward Measures

### Sharpe Ratio

\[SR = \frac{r - r_f}{\sigma}\]

The Sharpe ratio is a parametric measure of risk-reward: excess return per unit of volatility. When assuming `r_f = 0`, it measures actual returns per unit of risk.

```{r}
#| echo: false
#| warning: false
#| message: false

ret.xts <- xts::xts(
  result.opt$ret,
  order.by = result.opt$date
)

sr <- PerformanceAnalytics::SharpeRatio(ret.xts, Rf = 0, FUN = "StdDev")

dplyr::tibble(
  Metric = base::rownames(sr),
  Value  = base::as.numeric(sr)
) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "8.8.1 Sharpe Ratio — Optimal Parameters (Training)",
    subtitle = "Multi-pair equal-weight portfolio | Rf = 0"
  ) %>%
  gt::fmt_number(columns = Value, decimals = 4)
```

### Omega Ratio

The Omega ratio measures the ratio of gains above a threshold to losses below it. `Omega > 1` means more upside than downside. Omega also considers the effect of fat tails, whereas Sharpe just assumes that returns are normally distributed.

```{r}
#| echo: false
#| warning: false
#| message: false

omega.val <- PerformanceAnalytics::Omega(ret.xts, L = 0)
omegarsr  <- PerformanceAnalytics::OmegaSharpeRatio(ret.xts)

# Return distribution
ret.vec <- base::as.numeric(ret.xts[ret.xts != 0])

p.omega <- dplyr::tibble(ret = ret.vec) %>%
  ggplot2::ggplot(ggplot2::aes(x = ret)) +
  ggplot2::geom_histogram(
    col  = "black",
    fill = "blue",
    ggplot2::aes(y = ggplot2::after_stat(density)),
    bins = 40
  ) +
  ggplot2::stat_function(
    fun  = stats::dnorm,
    args = list(
      mean = base::mean(ret.vec, na.rm = TRUE),
      sd   = stats::sd(ret.vec, na.rm = TRUE)
    ),
    col  = "red",
    linewidth = 1
  ) +
  ggplot2::geom_vline(xintercept = 0, col = "black") +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title    = paste0(
      "8.8.2 Sharpe vs Omega | Omega = ", base::round(base::as.numeric(omega.val), 3),
      " | Omega-Sharpe = ", base::round(base::as.numeric(omegarsr), 3)
    ),
    subtitle = "Multi-pair portfolio daily returns | Blue bars = empirical | Red curve = normal",
    x        = "Daily Return",
    y        = "Density"
  )

p.omega
```

Comparing Omega to Sharpe, we can see that our strategy has more return centered around 0, and has a fatter tails. This is why the two ratios diverge.

### Full Performance

```{r}
#| echo: false
#| warning: false
#| message: false

stats.train <- result.opt %>%
  dplyr::select(date, ret) %>%
  RTL::tradeStats()

dplyr::as_tibble(stats.train) %>%
  tidyr::pivot_longer(
    cols      = dplyr::everything(),
    names_to  = "Metric",
    values_to = "Value"
  ) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "8.4.5 Performance Assessment — Training Period",
    subtitle = paste0(
      "Multi-pair equal-weight portfolio | Monthly rebalancing",
      " | z.entry = ", optimal$z.entry,
      " | n.ma = ", optimal$n.ma
    )
  ) %>%
  gt::fmt_number(columns = Value, decimals = 4)
```

### Drawdown Analysis

Drawdowns measure the percentage loss since the prior peak. Max Drawdown Length indicates how long the strategy has been underwater.

```{r}
#| echo: false
#| warning: false
#| message: false

# Drawdown table (top 10 worst)
dd.table <- PerformanceAnalytics::table.Drawdowns(ret.xts, top = 10)

dd.table %>%
  dplyr::as_tibble() %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "8.8.6 Top 10 Drawdowns — Training Period",
    subtitle = "Multi-pair equal-weight portfolio | Optimal parameters"
  ) %>%
  gt::fmt_number(columns = Depth, decimals = 4)
```

```{r}
#| echo: false
#| warning: false
#| message: false

# Drawdown chart (ggplot2 equivalent of PerformanceAnalytics::chart.Drawdown)
cumeq.vec <- result.opt$cumeq
running.max <- base::cummax(cumeq.vec)
drawdown    <- cumeq.vec / running.max - 1

dplyr::tibble(
  date     = result.opt$date,
  drawdown = drawdown
) %>%
  tidyr::drop_na() %>%
  ggplot2::ggplot(ggplot2::aes(x = date, y = drawdown)) +
  ggplot2::geom_area(fill = "red", alpha = 0.4) +
  ggplot2::geom_line(col = "red") +
  ggplot2::geom_hline(yintercept = 0, col = "black") +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title    = "8.8.6–8.8.7 Drawdown — Multi-pair Portfolio",
    subtitle = "Percentage loss from prior equity peak | Training period",
    x        = "",
    y        = "Drawdown"
  )
```

### Correlation to Underlying (SPY Benchmark)

When correlation ρ = 1, the strategy returns equal long buy-and-hold. When ρ = −1, it equals short buy-and-hold. A market-neutral pairs strategy should have ρ near zero, confirming it does not simply replicate a directional SPY position.

```{r}
#| echo: false
#| warning: false
#| message: false

# SPY returns for benchmark comparison
spy.ret <- etf.ret.train %>%
  dplyr::filter(symbol == "SPY") %>%
  dplyr::select(date, ret.spy = ret)

strat.ret <- result.opt %>%
  dplyr::select(date, ret) %>%
  tidyr::drop_na()

df.corr <- strat.ret %>%
  dplyr::inner_join(spy.ret, by = "date")

cor.val <- stats::cor(df.corr$ret, df.corr$ret.spy, use = "complete.obs")

p.corr <- df.corr %>%
  ggplot2::ggplot(ggplot2::aes(x = ret.spy, y = ret)) +
  ggplot2::geom_point(col = "grey20", alpha = 0.3, size = 0.8) +
  ggplot2::geom_smooth(method = "lm", col = "red", se = TRUE) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title    = paste0(
      "8.8.8 Portfolio vs SPY Correlation | ρ = ",
      base::round(cor.val, 3)
    ),
    subtitle = "Near-zero correlation confirms market neutrality | Red = linear fit",
    x        = "SPY Daily Return",
    y        = "Multi-pair Portfolio Return"
  )

p.corr
```

------------------------------------------------------------------------

## Out-of-Sample Evaluation

Parameters selected in from the Optimization section are applied to the **test period (Jan 2025 – Mar 2026)** without any refitting. All model parameters are fit exclusively on training data, i.e. the test period is never touched until final evaluation.

```{r}
#| echo: false
#| warning: false
#| message: false

result.test <- portfolio_pairs(
  pairs.df    = pairs.tradeable,
  prices.wide = log.prices.wide.test,
  open.wide   = log.open.wide.test,
  n.ma        = optimal$n.ma,
  z.entry     = optimal$z.entry,
  z.stop      = 3
)

stats.test <- result.test %>%
  dplyr::select(date, ret) %>%
  RTL::tradeStats()
```

```{r}
#| echo: false
#| warning: false
#| message: false

# Train vs Test comparison table
dplyr::bind_rows(
  dplyr::as_tibble(stats.train) %>%
    tidyr::pivot_longer(dplyr::everything(), names_to = "Metric", values_to = "Train"),
  dplyr::as_tibble(stats.test) %>%
    tidyr::pivot_longer(dplyr::everything(), names_to = "Metric", values_to = "Test")
) %>%
  dplyr::group_by(Metric) %>%
  dplyr::summarise(
    Train = dplyr::first(Train),
    Test  = dplyr::last(Test),
    .groups = "drop"
  ) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "8.10 Train vs Out-of-Sample Performance",
    subtitle = paste0(
      "Multi-pair equal-weight portfolio | Monthly rebalancing",
      " | z.entry = ", optimal$z.entry,
      " | n.ma = ", optimal$n.ma
    )
  ) %>%
  gt::fmt_number(columns = c(Train, Test), decimals = 4)
```

```{r}
#| echo: false
#| warning: false
#| message: false

# Out-of-sample portfolio equity curve
result.test %>%
  tidyr::drop_na(cumeq) %>%
  ggplot2::ggplot(ggplot2::aes(x = date, y = cumeq)) +
  ggplot2::geom_line(col = "blue", linewidth = 0.8) +
  ggplot2::geom_hline(yintercept = 1, col = "grey20") +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title    = "8.10 Out-of-Sample Portfolio Equity",
    subtitle = paste0(
      "Jan 2025 – Mar 2026 | Parameters selected on training data only",
      " | z.entry = ", optimal$z.entry, " | n.ma = ", optimal$n.ma
    ),
    x = "",
    y = "Cumulative Equity"
  )
```

### Interpretation - Out-of-Sample Results

**What the numbers say:**

-   **Train Sharpe 0.71, Test Sharpe 0.60:** The strategy earns positive risk-adjusted returns out of sample. The degradation is modest, not a collapse, which is the best realistic outcome in quantitative trading.

-   **Train CumReturn 2.41, Test CumReturn 0.05:** This comparison is misleading on its face. Train covers 24 years; test covers 15 months. The test window produced approximately 4.5% gross return in 15 months from a market-neutral strategy with no directional exposure, which is consistent with an annualized return of 3.7%.

-   **Train vs Test DD.Max:** The test drawdown of 4.25% is shallower than the train drawdown of 8.98%, but this is partly a time artifact. Fifteen months provides fewer opportunities for a large drawdown to develop. The equity curve confirms the portfolio recovered cleanly after the mid-2025 dip, which is the behavioural signature of genuine mean reversion.

-   **Win rate and time-in-market are stable:** %.Win and %.InMrkt are nearly identical across train and test. The signal generation logic is consistent, not a product of overfitting to a specific training regime.

**Why some degradation occurs:**

-   **Regime change:** The cointegration relationships were estimated on 24 years of data. The 2025-2026 period introduced tariff shocks, rate cuts, and sector rotation that may shift spread dynamics across multiple pairs simultaneously.

-   **Slow reversion:** Sector ETF pairs mean-revert on timescales measured in months. Fifteen months of test data contains very few complete half-life cycles per pair, too few to distinguish a genuine edge from noise with statistical confidence.

-   **Parameter fit:** Even with walk-forward discipline and plateau selection, the optimal parameters were identified on a 24-year training regime. Some of that fit will not transfer perfectly to a regime the optimizer never saw.

**What this means for the strategy:**

The out-of-sample result does not invalidate the strategy. A Sharpe of 0.60 and an annualized return of 3.7% with a max drawdown under 5% is a commercially viable outcome for a market-neutral strategy. The multi-pair portfolio structure reduces the single-pair concentration risk present in simpler implementations, and the equity curve shows the portfolio ending the test period above its starting value despite a difficult macro environment. A definitive verdict requires 3 to 5 years of out-of-sample data covering multiple complete business cycles, but the 15-month result provides no evidence that the strategy edge has broken down.

------------------------------------------------------------------------

## Literature Flow — Which Paper Supports Which Step

Each step in the strategy pipeline is grounded in academic literature. The diagram below traces the intellectual lineage: which paper motivated which modeling choice, and where the class notes provide direct instruction versus where external research was required.

```{mermaid}
graph LR
    subgraph PAPERS["<b>Academic Sources</b>"]
        P1["<b>Cartea, Cucuringu & Jin (2023)</b><br/>SSRN 4560455<br/>Correlation-matrix clustering<br/>→ market-neutral portfolios"]
        P2["<b>Yang, Huang & Chen (2023)</b><br/>Hierarchical clustering +<br/>Kalman filter blueprint"]
        P3["<b>Simonson (2018)</b><br/>OU process for ETF spreads<br/>University of Washington"]
        P4["<b>Avellaneda & Lee (2008)</b><br/>SSRN 1153505<br/>Statistical arbitrage<br/>Sharpe 1.44–1.51"]
        P5["<b>Engle & Granger (1987)</b><br/>Nobel Prize — cointegration<br/>ADF residual test"]
        P6["<b>Kalman (1960)</b><br/>State-space filtering<br/>Optimal linear estimator"]
    end

    subgraph CLASS["<b>FIN 452 Class Notes</b>"]
        C1["Module 2<br/>Trading Process"]
        C2["Module 3<br/>Sector Rotation"]
        C3["Module 8<br/>Quantitative Trading"]
    end

    subgraph STEPS["<b>Strategy Steps</b>"]
        S1["Step 1: Clustering"]
        S2["Step 2: Cointegration"]
        S3["Step 3: OU Half-Life"]
        S4["Step 4: Kalman Filter"]
        S5["8.4 Strategy Function"]
        S6["8.7 Optimization"]
        S7["8.8 Risk / Reward"]
    end

    P1 --> S1
    P2 --> S1
    P2 --> S2
    P2 --> S4
    P3 --> S3
    P4 --> S4
    P5 --> S2
    P6 --> S4
    C1 --> S5
    C2 --> S1
    C3 --> S5
    C3 --> S6
    C3 --> S7
```

```{r}
#| echo: false
#| warning: false
#| message: false

dplyr::tibble(
  `Strategy Step` = c(
    "Step 1 — Clustering",
    "Step 1 — Clustering",
    "Step 2 — Cointegration",
    "Step 2 — Cointegration",
    "Step 3 — OU Half-Life",
    "Step 4 — Kalman Filter",
    "Step 4 — Kalman Filter",
    "Step 4 — Kalman Filter"
  ),
  Source = c(
    "Cartea, Cucuringu & Jin (2023)",
    "Yang, Huang & Chen (2023)",
    "Engle & Granger (1987)",
    "Yang, Huang & Chen (2023)",
    "Simonson (2018)",
    "Kalman (1960)",
    "Avellaneda & Lee (2008)",
    "Yang, Huang & Chen (2023)"
  ),
  `Key Contribution` = c(
    "Clustering on correlation matrix → mean-reverting portfolios; Sharpe > 1 demonstrated on equities",
    "Hierarchical clustering as Layer 1 before cointegration + Kalman; direct blueprint for this strategy",
    "Cointegration theory and the ADF test for residual stationarity — Nobel Prize foundational work",
    "Cointegration as the statistical gate between clustering and Kalman — 'the two steps are complements'",
    "OU process calibration for ETF pairs; half-life as holding-period filter; sector-ETF focus",
    "Original state-space filtering paper; optimal linear estimator for time-varying hedge ratio",
    "Full statistical arbitrage framework on U.S. equities; Sharpe 1.44–1.51; historical benchmark",
    "End-to-end workflow: clustering → cointegration → Kalman → trading signal in a single paper"
  )
) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Literature Map — Academic and Class Sources by Strategy Step",
    subtitle = "Each row traces how a specific paper or class note motivates a modeling choice"
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_body(columns = `Strategy Step`)
  )
```

**How the papers connect:** Yang, Huang & Chen (2023) is the closest single-paper blueprint — it uses hierarchical clustering as a pre-filter, cointegration as a validation gate, and a Kalman filter for dynamic hedge ratios, exactly as implemented here. Cartea et al. (2023) provides the theoretical justification for *why* correlation-matrix clustering produces mean-reverting portfolios. Simonson (2018) fills the OU half-life gap not covered in class notes. Avellaneda & Lee (2008) is the historical anchor — it demonstrates that statistical arbitrage on U.S. equities can achieve Sharpe \> 1.4, setting the performance benchmark. Engle & Granger (1987) and Kalman (1960) are foundational — they provide the econometric and filtering tools that make the strategy mathematically rigorous.

------------------------------------------------------------------------

## Weaknesses and Limitations

No backtested strategy should be presented without an honest assessment of what could go wrong. The weaknesses below are organized from most to least severe.

```{r}
#| echo: false
#| warning: false
#| message: false

dplyr::tibble(
  `#` = 1:10,
  Weakness = c(
    "Single-pair concentration (addressed)",
    "Slow mean-reversion",
    "No transaction costs",
    "Weak cluster separation",
    "Overfitting risk in optimization",
    "Small, homogeneous universe",
    "Fixed train/test split",
    "No regime detection",
    "Look-ahead in data construction",
    "Short out-of-sample window"
  ),
  Severity = c(
    "Low", "High", "Medium", "Medium", "Medium",
    "Medium", "Low", "Low", "Low", "Low"
  ),
  Description = c(
    "Addressed: all tradeable pairs are traded simultaneously with equal weight and monthly rebalancing. If one pair weakens, the others continue contributing. Residual risk is that all pairs could weaken simultaneously in an extreme regime shift.",
    "Sector ETF pairs mean-revert on timescales measured in months. The strategy generates very few round-trip trades per year per pair. In the 15-month test window there may be only 2 to 4 complete cycles across the portfolio — far too few for statistical confidence. The strategy is structurally underpowered for short evaluation horizons.",
    "Bid/ask spreads, commissions, and slippage are not modeled. Sector ETFs have tight spreads (~0.01–0.02%), so the impact is small for low-frequency trading, but it is non-zero and compounds over 24 years of training data.",
    "The K-Means silhouette score (~0.16) is well below the 0.5 'strong' threshold. All 10 ETFs share dominant systematic exposure to SPY, making cluster boundaries inherently fuzzy. Clustering serves an organizational role, not a statistical one — cointegration is the true filter.",
    "The 120-point parameter grid was optimized on 24 years of training data. Even with the plateau robustness method, there is no guarantee that the selected parameters generalize to future regimes. The composite Z-score mitigates but does not eliminate this risk.",
    "Ten sector ETFs drawn from one index (S&P 500) offer limited diversity. All sectors are U.S. large-cap, making the universe structurally correlated. A broader universe (international ETFs, factor ETFs, commodities) would provide more independent pair candidates.",
    "A single 80/20 train/test split at Jan 2025 does not test performance across multiple regimes. Rolling-window or expanding-window walk-forward validation would be more rigorous but computationally expensive for this assignment scope.",
    "The Kalman filter adapts the hedge ratio continuously, but the strategy has no explicit mechanism to detect when the cointegration relationship has structurally broken. A regime-switching model (Markov, CUSUM) could flag breakdown periods.",
    "The train/test split date is known a priori. While no test-period information leaks into model fitting, the choice of split date is itself a decision that could be optimized. This is a subtle form of look-ahead.",
    "Fifteen months (Jan 2025 – Mar 2026) contains ~3 half-life cycles at best. Any conclusion about out-of-sample performance — positive or negative — has very wide confidence intervals."
  )
) %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Weaknesses and Limitations",
    subtitle = "Honest assessment of what could go wrong and what was not modeled"
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(color = "red", weight = "bold"),
    locations = gt::cells_body(columns = Severity, rows = Severity == "High")
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(color = "orange", weight = "bold"),
    locations = gt::cells_body(columns = Severity, rows = Severity == "Medium")
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(color = "blue"),
    locations = gt::cells_body(columns = Severity, rows = Severity == "Low")
  )
```

**Key takeaway:** Single-pair concentration has been addressed by trading all tradeable pairs simultaneously with equal weight and monthly rebalancing. The remaining highest-severity weakness is slow reversion, which is intrinsic to the sector ETF universe. The practical implication is that this strategy requires a multi-year evaluation window to assess properly, and the 15-month test period in this assignment is insufficient for statistical significance.

------------------------------------------------------------------------

## Conclusion

This project implements the full quantitative pairs trading workflow: starting from the hypothesis that sector ETFs share systematic risk factors that create exploitable mean-reverting spreads, then filtering the universe through a disciplined four-step pipeline (K-Means Clustering, Engle-Granger Cointegration, Ornstein-Uhlenbeck Half-Life, and Kalman-Filtered Dynamic Hedging) to identify tradeable pairs with statistically validated mean-reversion properties.

Rather than betting on a single pair, the strategy trades all surviving pairs simultaneously in an equal-weight portfolio with monthly rebalancing. This design choice directly addresses the most dangerous weakness in traditional pairs trading: single-pair concentration risk. When one pair's spread diverges permanently, the other pairs continue generating returns, and the monthly rebalancing prevents any one pair from dominating portfolio exposure.

The parameter selection process goes beyond naive in-sample optimization. The 3D Sharpe surface (8.9) revealed that strategy performance is not a single peak but a broad plateau across a range of z.entry and n.ma combinations. The plateau robustness method (8.9.2) selects parameters from the centre of this stable region rather than the fragile maximum, producing parameters that are more likely to survive regime change.

Out of sample, the strategy delivered a Sharpe ratio of 0.60 and an annualized return of 3.7% with a maximum drawdown under 5% over the January 2025 to March 2026 test period. These results represent modest but real degradation from the training period, which is the expected outcome for any honestly evaluated quantitative strategy. The 15-month test window is too short to deliver a definitive verdict on a slow mean-reversion strategy, but the results show no evidence of structural breakdown.

------------------------------------------------------------------------

## Lessons Learned
